---
title: "Simulation Study in Section 4.7"
output:
  html_document:
    code_folding: show
    highlight: textmate
    df_print: paged
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Common code

```{r setup, message=FALSE, warning=FALSE}
library(ordinal)
library(VGAM)
library(CUB)
library(zipfR)
library(dplyr)
library(ggplot2)
library(patchwork)

# This R Markdown file and common_functions.R should be in the same directory.
source(file.path("code", "common_functions.R"))

n_obs <- 500L
n_rep <- 500L
m_cat <- 4L
base_seed <- 10000L
```

## Data-generating mechanisms

```{r data-generation-functions}
# True conditional survival probabilities under the proportional odds model.
true_Q_pom <- function(eta, m = 4L,
                       alpha = c(-1.5, -0.2, 1.5)) {
  stopifnot(length(alpha) == m - 1L)

  Q <- outer(
    eta,
    alpha,
    FUN = function(eta_value, alpha_value) {
      plogis(eta_value - alpha_value)
    }
  )

  colnames(Q) <- paste0("Q", seq_len(m - 1L))
  Q
}


# True conditional survival probabilities under S2{rho_1,1(x; j)}.
true_Q_s2_rho11 <- function(eta, m = 4L) {
  j_grid <- seq_len(m - 1L)

  Q <- sapply(j_grid, function(j) {
    (1 - (j / m)^exp(eta))^2
  })

  colnames(Q) <- paste0("Q", j_grid)
  Q
}


# True conditional survival probabilities under H2{rho_1,1(x; j)} with k = 1.5.
true_Q_h2_rho11 <- function(eta, m = 4L, k = 1.5) {
  j_grid <- seq_len(m - 1L)

  Q <- sapply(j_grid, function(j) {
    (1 - (j / m)^exp(eta))^k
  })

  colnames(Q) <- paste0("Q", j_grid)
  Q
}


# Generate one ordinal data set from a supplied conditional survival function.
generate_ordinal_data <- function(n, m, beta, true_Q_function, seed) {
  set.seed(seed)

  x1 <- runif(n, min = -1, max = 1)
  x2 <- rbinom(n, size = 1, prob = 0.5)
  eta <- beta[1] * x1 + beta[2] * x2

  Q_true <- true_Q_function(eta = eta, m = m)
  F_true <- 1 - Q_true

  u <- runif(n)
  y <- 1L + rowSums(
    matrix(u, nrow = n, ncol = m - 1L) > F_true
  )

  dat <- data.frame(
    y = factor(y, levels = seq_len(m), ordered = TRUE),
    x1 = x1,
    x2 = x2,
    eta = eta
  )

  list(data = dat, Q_true = Q_true)
}
```

## Prediction and evaluation functions

```{r prediction-functions}
# Convert category probabilities to conditional survival probabilities.
prob_to_Q <- function(prob) {
  prob <- as.matrix(prob)
  m <- ncol(prob)

  Q <- sapply(seq_len(m - 1L), function(j) {
    rowSums(prob[, (j + 1L):m, drop = FALSE])
  })

  Q <- as.matrix(Q)
  colnames(Q) <- paste0("Q", seq_len(m - 1L))
  Q
}


# Mean squared error of the fitted conditional survival probabilities.
calc_MSE_Q <- function(Q_hat, Q_true) {
  mean((Q_hat - Q_true)^2)
}


# Fitted conditional survival probabilities from a cumulative link model.
get_Q_clm <- function(model, dat) {
  newdat <- dat[, c("x1", "x2"), drop = FALSE]
  prob <- predict(model, newdata = newdat, type = "prob")$fit
  prob_to_Q(prob)
}


# Fitted conditional survival probabilities from a sequential model.
get_Q_seq <- function(model, dat) {
  prob <- predict(model, newdata = dat, type = "response")
  prob_to_Q(prob)
}


# Fitted conditional survival probabilities from a CUB model.
get_Q_cub <- function(model, dat, m) {
  par_hat <- as.numeric(coef(model))
  pai_hat <- par_hat[1]
  gama_hat <- par_hat[-1]

  W <- as.matrix(dat[, c("x1", "x2")])
  n <- nrow(dat)

  prob <- sapply(seq_len(m), function(r) {
    CUB::probcub0q(
      m = m,
      ordinal = rep(r, n),
      W = W,
      pai = pai_hat,
      gama = gama_hat
    )
  })

  prob_to_Q(prob)
}


# Fitted conditional survival probabilities from a GCLM.
get_Q_gclm <- function(fit, X, m,
                       surv_candidates,
                       rhoj_candidates) {
  eta <- drop(X %*% fit$beta)
  eta_value <- geta_candidates[[fit$geta]]$func(eta)

  Q <- sapply(seq_len(m - 1L), function(j) {
    t_j <- do.call(
      rhoj_candidates[[fit$rhoj]]$fun,
      c(
        list(
          eta_val = eta_value,
          j = j,
          m = m
        ),
        as.list(fit$rhoj_par)
      )
    )

    do.call(
      surv_candidates[[fit$surv]]$func,
      c(
        list(t = t_j),
        as.list(fit$surv_par)
      )
    )
  })

  Q <- as.matrix(Q)
  colnames(Q) <- paste0("Q", seq_len(m - 1L))
  Q
}
```

## Model-fitting functions

```{r fitting-functions}
# The nonidentifiable combinations excluded under the continuous-covariate setting.
is_identifiable_gclm <- function(surv_name, rhoj_name) {
  !(surv_name == "H4" &&
      rhoj_name %in% c("rho_14", "rho_44", "rho_54"))
}


# Fit the cumulative, sequential, and CUB competitors in one replication.
fit_competing_models <- function(dat, Q_true, m, rep_id) {
  n <- nrow(dat)
  fit_rows <- list()

  clm_models <- list(
    Cum_Logit = ordinal::clm(
      y ~ x1 + x2,
      data = dat,
      link = "logit"
    ),
    Cum_Probit = ordinal::clm(
      y ~ x1 + x2,
      data = dat,
      link = "probit"
    ),
    Cum_Cloglog = ordinal::clm(
      y ~ x1 + x2,
      data = dat,
      link = "cloglog"
    )
  )

  for (model_name in names(clm_models)) {
    model <- clm_models[[model_name]]
    Q_hat <- get_Q_clm(model, dat)

    fit_rows[[model_name]] <- data.frame(
      rep = rep_id,
      model = model_name,
      rho_a = NA_integer_,
      surv = NA_character_,
      rhoj = NA_character_,
      logLik = as.numeric(logLik(model)),
      AIC = AIC(model),
      BIC = BIC(model),
      MSE_Q = calc_MSE_Q(Q_hat, Q_true)
    )
  }

  seq_model <- VGAM::vglm(
    y ~ x1 + x2,
    family = VGAM::sratio(parallel = TRUE),
    data = dat
  )

  Q_hat_seq <- get_Q_seq(seq_model, dat)

  fit_rows[["Seq_Logit"]] <- data.frame(
    rep = rep_id,
    model = "Seq_Logit",
    rho_a = NA_integer_,
    surv = NA_character_,
    rhoj = NA_character_,
    logLik = as.numeric(logLik(seq_model)),
    AIC = AIC(seq_model),
    BIC = BIC(seq_model),
    MSE_Q = calc_MSE_Q(Q_hat_seq, Q_true)
  )

  invisible(
    capture.output(
      cub_model <- CUB::GEM(
        Formula::Formula(y ~ 0 | x1 + x2 | 0),
        family = "cub",
        data = dat
      )
    )
  )

  Q_hat_cub <- get_Q_cub(cub_model, dat, m)
  cub_par <- cub_model$estimates
  cub_logLik <- cub_model$loglik

  fit_rows[["CUB"]] <- data.frame(
    rep = rep_id,
    model = "CUB",
    rho_a = NA_integer_,
    surv = NA_character_,
    rhoj = NA_character_,
    logLik = cub_logLik,
    AIC = -2 * cub_logLik + 2 * length(cub_par),
    BIC = -2 * cub_logLik + log(n) * length(cub_par),
    MSE_Q = calc_MSE_Q(Q_hat_cub, Q_true)
  )

  dplyr::bind_rows(fit_rows)
}


# Construct the GCLM candidate grid for a scenario.
make_gclm_grid <- function(scenario, surv_candidates, rhoj_candidates) {
  if (isTRUE(scenario$fit_all_gclms)) {
    grid <- expand.grid(
      surv = names(surv_candidates),
      rhoj = names(rhoj_candidates),
      stringsAsFactors = FALSE
    )

    keep <- mapply(
      is_identifiable_gclm,
      grid$surv,
      grid$rhoj
    )

    return(grid[keep, , drop = FALSE])
  }

  data.frame(
    surv = scenario$gclm_surv,
    rhoj = scenario$gclm_rhoj,
    stringsAsFactors = FALSE
  )
}


# Fit the GCLM candidates specified for one scenario.
fit_gclm_models <- function(dat, Q_true, m, rep_id, scenario) {
  n <- nrow(dat)
  X <- model.matrix(~ -1 + x1 + x2, data = dat)
  y_num <- as.numeric(dat$y)

  surv_candidates <- make_surv_func(m = m)
  rhoj_candidates <- make_rhoj_candidates()
  candidate_grid <- make_gclm_grid(
    scenario,
    surv_candidates,
    rhoj_candidates
  )

  all_results <- list()

  for (i in seq_len(nrow(candidate_grid))) {
    surv_name <- candidate_grid$surv[i]
    rhoj_name <- candidate_grid$rhoj[i]
    eta_name <- get_eta_name_for_rhoj(rhoj_name)

    surv_info <- surv_candidates[[surv_name]]
    eta_info <- geta_candidates[[eta_name]]
    rhoj_info <- rhoj_candidates[[rhoj_name]]

    init_fit <- get_data_driven_init(
      y = y_num,
      m = m,
      surv_info = surv_info,
      eta_info = eta_info,
      rhoj_info = rhoj_info
    )

    surv_info$init <- init_fit$surv_init
    rhoj_info$init <- init_fit$rhoj_init

    fit <- fit_one_combo(
      X = X,
      y = y_num,
      m = m,
      n = n,
      surv_info = surv_info,
      eta_info = eta_info,
      rhoj_info = rhoj_info,
      method = "BFGS"
    )

    key <- paste(surv_name, eta_name, rhoj_name, sep = "__")

    all_results[[key]] <- c(
      list(
        surv = surv_name,
        geta = eta_name,
        rhoj = rhoj_name
      ),
      fit
    )
  }

  best_by_rhoa <- select_best_by_rhoa(
    all_results = all_results,
    criterion = "BIC",
    require_convergence = TRUE
  )

  if (is.null(best_by_rhoa) || nrow(best_by_rhoa) == 0L) {
    stop("No converged GCLM was available in replication ", rep_id, ".")
  }

  fit_rows <- vector("list", nrow(best_by_rhoa))

  for (i in seq_len(nrow(best_by_rhoa))) {
    rho_a <- best_by_rhoa$rho_a[i]
    key <- best_by_rhoa$key[i]
    fit <- all_results[[key]]

    Q_hat <- get_Q_gclm(
      fit = fit,
      X = X,
      m = m,
      surv_candidates = surv_candidates,
      rhoj_candidates = rhoj_candidates
    )

    fit_rows[[i]] <- data.frame(
      rep = rep_id,
      model = paste0("GCLM_a", rho_a),
      rho_a = rho_a,
      surv = fit$surv,
      rhoj = fit$rhoj,
      logLik = -fit$nll,
      AIC = fit$aic,
      BIC = fit$bic,
      MSE_Q = calc_MSE_Q(Q_hat, Q_true)
    )
  }

  dplyr::bind_rows(fit_rows)
}


# Run one replication for a supplied scenario.
run_one_replication <- function(rep_id, scenario,
                                n = 500L, m = 4L,
                                seed = 10000L + rep_id) {
  simulated <- generate_ordinal_data(
    n = n,
    m = m,
    beta = scenario$beta,
    true_Q_function = scenario$true_Q_function,
    seed = seed
  )

  competitor_results <- fit_competing_models(
    dat = simulated$data,
    Q_true = simulated$Q_true,
    m = m,
    rep_id = rep_id
  )

  gclm_results <- fit_gclm_models(
    dat = simulated$data,
    Q_true = simulated$Q_true,
    m = m,
    rep_id = rep_id,
    scenario = scenario
  )

  dplyr::bind_rows(competitor_results, gclm_results)
}


# Run all replications for one scenario.
run_simulation <- function(scenario,
                           R = 500L,
                           n = 500L,
                           m = 4L,
                           seed_start = 10000L) {
  fit_list <- vector("list", R)
  failed_rep <- integer(0)

  for (r in seq_len(R)) {
    message(
      "Scenario ", scenario$id,
      ": replication ", r, " of ", R
    )

    result <- tryCatch(
      run_one_replication(
        rep_id = r,
        scenario = scenario,
        n = n,
        m = m,
        seed = seed_start + r
      ),
      error = function(e) {
        message(
          "Scenario ", scenario$id,
          ", replication ", r,
          " failed: ", conditionMessage(e)
        )
        NULL
      }
    )

    if (is.null(result)) {
      failed_rep <- c(failed_rep, r)
    } else {
      fit_list[[r]] <- result
    }
  }

  list(
    fit = dplyr::bind_rows(fit_list),
    failed_rep = failed_rep
  )
}
```

# Scenario (a): POM

Data are generated from a proportional odds model with
\(\boldsymbol{\alpha}=(-1.5,-0.2,1.5)^\top\) and
\(\boldsymbol{\beta}=(1,1)^\top\). All identifiable GCLM candidates are
fitted, and the candidate with the smallest BIC is used for comparison.

```{r scenario-a, warning=FALSE}
scenario_a <- list(
  id = "a",
  beta = c(1, 1),
  true_Q_function = true_Q_pom,
  fit_all_gclms = TRUE
)

results_a <- run_simulation(
  scenario = scenario_a,
  R = n_rep,
  n = n_obs,
  m = m_cat,
  seed_start = base_seed
)
```

# Scenario (b): GCLM with S2{rho_1,1(x; j)}

Data are generated from the GCLM defined by \(S_2\{\rho_{1,1}(\mathbf{x};j)\}\)
with \(\boldsymbol{\beta}=(0.7,0.7)^\top\). The corresponding GCLM is fitted.

```{r scenario-b, warning=FALSE}
scenario_b <- list(
  id = "b",
  beta = c(0.7, 0.7),
  true_Q_function = true_Q_s2_rho11,
  fit_all_gclms = FALSE,
  gclm_surv = "S2",
  gclm_rhoj = "rho_11"
)

results_b <- run_simulation(
  scenario = scenario_b,
  R = n_rep,
  n = n_obs,
  m = m_cat,
  seed_start = base_seed
)
```

# Scenario (c): GCLM with H2{rho_1,1(x; j)}

Data are generated from the GCLM defined by \(H_2\{\rho_{1,1}(\mathbf{x};j)\}\)
with \(\boldsymbol{\beta}=(0.7,0.7)^\top\) and \(k=1.5\). The corresponding
GCLM is fitted.

```{r scenario-c, warning=FALSE}
scenario_c <- list(
  id = "c",
  beta = c(0.7, 0.7),
  true_Q_function = true_Q_h2_rho11,
  fit_all_gclms = FALSE,
  gclm_surv = "H2",
  gclm_rhoj = "rho_11"
)

results_c <- run_simulation(
  scenario = scenario_c,
  R = n_rep,
  n = n_obs,
  m = m_cat,
  seed_start = 1221
)
```

# Figure 4

The result objects created above are used directly. No external RDS files are
required to generate Figure 4.

```{r figure-functions}
model_order <- c(
  "Cum_Logit",
  "Cum_Probit",
  "Cum_Cloglog",
  "Seq_Logit",
  "CUB",
  "GCLM"
)

model_labels <- c(
  "Cum_Logit" = "POM",
  "Cum_Probit" = "CLM\nprobit",
  "Cum_Cloglog" = "CLM\ncloglog",
  "Seq_Logit" = "Sequential",
  "CUB" = "CUB",
  "GCLM" = "GCLM"
)

table_labels <- c(
  "Cum_Logit" = "POM",
  "Cum_Probit" = "CLM probit",
  "Cum_Cloglog" = "CLM cloglog",
  "Seq_Logit" = "Sequential",
  "CUB" = "CUB",
  "GCLM" = "GCLM"
)

box_style <- list(
  width = 0.65,
  fill = "white",
  colour = "black",
  box.linewidth = 0.6,
  median.linewidth = 0.8,
  whisker.linetype = 2,
  whisker.linewidth = 0.5,
  staplewidth = 0.5,
  staple.linewidth = 0.5,
  outlier.shape = 1,
  outlier.size = 1.3,
  outlier.stroke = 0.45
)


# Select the overall BIC-best GCLM and combine it with the competitors.
prepare_plot_data <- function(simulation_result) {
  fit_results <- simulation_result$fit

  gclm_overall <- fit_results %>%
    filter(grepl("^GCLM_a", model)) %>%
    group_by(rep) %>%
    slice_min(BIC, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(model = "GCLM")

  comparison_results <- fit_results %>%
    filter(!grepl("^GCLM_a", model)) %>%
    bind_rows(gclm_overall) %>%
    mutate(
      model = factor(model, levels = model_order),
      deviance = -2 * logLik
    )

  comparison_results
}

# Calculate model-selection frequencies for AIC or BIC.
calculate_selection_frequency <- function(comparison_results,
                                          criterion = c("AIC", "BIC"),
                                          tolerance = 1e-8) {
  criterion <- match.arg(criterion)
  n_successful <- dplyr::n_distinct(comparison_results$rep)

  table_labels <- c(
    "Cum_Logit" = "POM",
    "Cum_Probit" = "CLM probit",
    "Cum_Cloglog" = "CLM cloglog",
    "Seq_Logit" = "Sequential",
    "CUB" = "CUB",
    "GCLM" = "GCLM"
  )

  comparison_results %>%
    group_by(rep) %>%
    mutate(
      best_value = min(.data[[criterion]], na.rm = TRUE),
      is_best = abs(.data[[criterion]] - best_value) < tolerance
    ) %>%
    filter(is_best) %>%
    summarise(
      winner = paste(
        table_labels[as.character(model)],
        collapse = " / "
      ),
      .groups = "drop"
    ) %>%
    count(winner, name = "Frequency") %>%
    mutate(Percent = 100 * Frequency / n_successful) %>%
    arrange(desc(Frequency), winner)
}

# Calculate the medians represented by the boxplot center lines.
calculate_boxplot_medians <- function(comparison_results) {
  comparison_results %>%
    group_by(model) %>%
    summarise(
      median_deviance = median(deviance, na.rm = TRUE),
      median_MSE = median(MSE_Q, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      model = table_labels[as.character(model)]
    ) %>%
    as.data.frame()
}


# Create the deviance and MSE panels for one scenario.
make_plot_pair <- function(comparison_results) {
  common_theme <- theme_bw(base_size = 12) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(
        colour = "black",
        fill = NA,
        linewidth = 0.4
      ),
      axis.text.x = element_text(
        angle = 0,
        hjust = 0.5,
        vjust = 1,
        size = 10
      ),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 12)
    )

  p_deviance <- ggplot(
    comparison_results,
    aes(x = model, y = deviance)
  ) +
    do.call(geom_boxplot, box_style) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Deviance") +
    common_theme

  p_mse <- ggplot(
    comparison_results,
    aes(x = model, y = MSE_Q)
  ) +
    do.call(geom_boxplot, box_style) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "MSE") +
    common_theme

  list(deviance = p_deviance, mse = p_mse)
}
```

## AIC and BIC selection frequencies

### Scenario (a)

```{r selection-frequencies-a}
comparison_a <- prepare_plot_data(results_a)

AIC_frequency_a <- calculate_selection_frequency(
  comparison_a,
  criterion = "AIC"
)

BIC_frequency_a <- calculate_selection_frequency(
  comparison_a,
  criterion = "BIC"
)

AIC_frequency_a
BIC_frequency_a
```

### Scenario (b)

```{r selection-frequencies-b}
comparison_b <- prepare_plot_data(results_b)

AIC_frequency_b <- calculate_selection_frequency(
  comparison_b,
  criterion = "AIC"
)

BIC_frequency_b <- calculate_selection_frequency(
  comparison_b,
  criterion = "BIC"
)

AIC_frequency_b
BIC_frequency_b
```

### Scenario (c)

```{r selection-frequencies-c}
comparison_c <- prepare_plot_data(results_c)

AIC_frequency_c <- calculate_selection_frequency(
  comparison_c,
  criterion = "AIC"
)

BIC_frequency_c <- calculate_selection_frequency(
  comparison_c,
  criterion = "BIC"
)

AIC_frequency_c
BIC_frequency_c
```

## Boxplot medians

### Scenario (a)

```{r boxplot-medians-a}
median_summary_a <- calculate_boxplot_medians(comparison_a)
median_summary_a
```

### Scenario (b)

```{r boxplot-medians-b}
median_summary_b <- calculate_boxplot_medians(comparison_b)
median_summary_b
```

### Scenario (c)

```{r boxplot-medians-c}
median_summary_c <- calculate_boxplot_medians(comparison_c)
median_summary_c
```

```{r figure-4, fig.width=14, fig.height=8}
plot_data_a <- comparison_a
plot_data_b <- comparison_b
plot_data_c <- comparison_c

plots_a <- make_plot_pair(plot_data_a)
plots_b <- make_plot_pair(plot_data_b)
plots_c <- make_plot_pair(plot_data_c)

p_deviance_a <- plots_a$deviance +
  labs(title = "(a) POM") +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 14
    )
  )

p_deviance_b <- plots_b$deviance +
  labs(
    title = expression(
      bold("(b) GCLM with") ~ S[2](rho[1*','*1](x*';'*j))
    )
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 14
    )
  )

p_deviance_c <- plots_c$deviance +
  labs(
    title = expression(
      bold("(c) GCLM with") ~ H[2](rho[1*','*1](x*';'*j))
    )
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 14
    )
  )

figure_4 <-
  (p_deviance_a | p_deviance_b | p_deviance_c) /
  (plots_a$mse | plots_b$mse | plots_c$mse) +
  plot_layout(
    widths = c(1, 1, 1),
    heights = c(1, 1)
  )

figure_4
```
