---
title: "Section 5.2: Retinopathy data (Table 6)"
output: html_document
editor_options:
  chunk_output_type: console
---

This file reproduces the numerical entries in Table 6. Knitting is not
required. Run the chunks from top to bottom; the final object
`table6_results` is printed as an ordinary R data frame.

## Setup and data

```{r setup, message=FALSE, warning=FALSE}
library(VGAM)
library(ordDisp)
library(catdata)
library(CUB)
library(zipfR)

source(file.path("code", "common_functions.R"))

required_interface <- c(
  "make_surv_func", "make_rhoj_candidates", "get_eta_name_for_rhoj",
  "get_data_driven_init", "fit_one_combo", "select_best_by_rhoa",
  "geta_candidates"
)

missing_interface <- required_interface[
  !vapply(required_interface, exists, logical(1), inherits = TRUE)
]

if (length(missing_interface) > 0L) {
  stop(
    "The following functions or objects are missing from common_functions.R: ",
    paste(missing_interface, collapse = ", ")
  )
}

if (!("T_np" %in% names(formals(fit_one_combo)))) {
  stop(
    "common_functions.R must contain the category-effect version of ",
    "fit_one_combo() with a T_np argument."
  )
}

data("retinopathy", package = "catdata")

retino_raw <- retinopathy[, c("RET", "SM", "DIAB", "GH", "BP")]
retino_raw$RET <- factor(
  retino_raw$RET,
  levels = c("0", "1", "2"),
  ordered = TRUE
)

if (anyNA(retino_raw)) {
  stop("The analysis variables contain missing values.")
}

retino_raw$.row_id <- seq_len(nrow(retino_raw))
continuous_vars <- c("DIAB", "GH", "BP")

# The complete-data estimates in the upper part of Table 6 use scale().
retino_full <- retino_raw
retino_full[continuous_vars] <- scale(retino_full[continuous_vars])
```

## Shared utilities

```{r shared-utilities}
format_estimate <- function(estimate, std_error, digits = 3L) {
  if (!is.finite(estimate) || !is.finite(std_error)) return(NA_character_)
  sprintf(paste0("%.", digits, "f (%.", digits, "f)"), estimate, std_error)
}

format_number <- function(x, digits = 3L) {
  if (!is.finite(x)) return(NA_character_)
  sprintf(paste0("%.", digits, "f"), x)
}

model_df <- function(fit) {
  ll <- stats::logLik(fit)
  df <- attr(ll, "df")
  if (is.null(df)) length(stats::coef(fit)) else as.integer(df)
}

make_stratified_folds <- function(y, v = 10L, repeats = 10L,
                                  seed = 1234L) {
  y <- droplevels(factor(y))

  if (any(table(y) < v)) {
    stop("Each response category must contain at least v observations.")
  }

  set.seed(seed)

  out <- lapply(seq_len(repeats), function(repeat_id) {
    fold_id <- integer(length(y))

    for (level in levels(y)) {
      index <- which(y == level)
      labels <- rep(seq_len(v), length.out = length(index))
      fold_id[index] <- sample(labels, length(index), replace = FALSE)
    }

    data.frame(
      repeat_id = repeat_id,
      fold_id = fold_id,
      row_id = seq_along(y)
    )
  })

  do.call(rbind, out)
}

fit_preprocessor <- function(train, continuous_vars) {
  center <- vapply(train[continuous_vars], mean, numeric(1))
  scale <- vapply(train[continuous_vars], stats::sd, numeric(1))

  if (any(!is.finite(center)) || any(!is.finite(scale)) || any(scale <= 0)) {
    stop("Invalid training-fold means or standard deviations.")
  }

  list(continuous_vars = continuous_vars, center = center, scale = scale)
}

apply_preprocessor <- function(data, preprocessor) {
  out <- data
  variables <- preprocessor$continuous_vars

  out[variables] <- Map(
    function(x, center, scale) (x - center) / scale,
    out[variables],
    preprocessor$center[variables],
    preprocessor$scale[variables]
  )

  out
}

prepare_prob <- function(prob, outcome_levels, tolerance = 1e-7) {
  prob <- as.matrix(prob)
  outcome_levels <- as.character(outcome_levels)

  if (ncol(prob) != length(outcome_levels)) {
    stop("The predicted probability matrix has the wrong number of columns.")
  }

  if (!is.null(colnames(prob)) &&
      all(outcome_levels %in% colnames(prob))) {
    prob <- prob[, outcome_levels, drop = FALSE]
  } else {
    colnames(prob) <- outcome_levels
  }

  if (any(!is.finite(prob))) {
    stop("Non-finite predicted probabilities were returned.")
  }

  if (any(prob < -tolerance) || any(prob > 1 + tolerance)) {
    stop("Predicted probabilities lie outside [0, 1].")
  }

  prob[prob < 0] <- 0
  prob[prob > 1] <- 1

  row_total <- rowSums(prob)
  if (any(abs(row_total - 1) > tolerance)) {
    stop("Predicted category probabilities do not sum to one.")
  }

  prob <- prob / row_total
  colnames(prob) <- outcome_levels
  prob
}

observation_scores <- function(truth, prob, outcome_levels) {
  truth <- factor(truth, levels = outcome_levels, ordered = TRUE)
  y <- as.integer(truth)
  n <- length(y)
  m <- length(outcome_levels)

  observed_probability <- prob[cbind(seq_len(n), y)]
  log_score <- -log(observed_probability)

  cumulative_probability <- t(apply(prob, 1L, cumsum))[
    , seq_len(m - 1L), drop = FALSE
  ]
  observed_cumulative <- outer(y, seq_len(m - 1L), FUN = "<=")
  ranked_score <- rowSums(
    (cumulative_probability - observed_cumulative)^2
  )

  data.frame(LogS = log_score, RPS = ranked_score)
}
```

## PPOM utilities

```{r ppom-utilities}
make_ppom_family <- function(nonparallel) {
  if (length(nonparallel) == 0L) {
    return(VGAM::cumulative(
      link = "logitlink", parallel = TRUE, reverse = FALSE
    ))
  }

  parallel_formula <- stats::as.formula(
    paste("FALSE ~", paste(nonparallel, collapse = " + "))
  )

  VGAM::cumulative(
    link = "logitlink",
    parallel = parallel_formula,
    reverse = FALSE
  )
}

fit_ppom_model <- function(formula, data, nonparallel,
                           etastart = NULL) {
  arguments <- list(
    formula = formula,
    family = make_ppom_family(nonparallel),
    data = data,
    trace = FALSE,
    na.action = stats::na.fail
  )

  if (!is.null(etastart)) arguments$etastart <- etastart
  do.call(VGAM::vglm, arguments)
}

extract_vgam_lrt_p <- function(lrt) {
  body <- lrt@Body
  p_column <- grep("^Pr\\(", colnames(body), value = TRUE)

  if (length(p_column) == 0L) {
    stop("No p-value column was found in the VGAM likelihood-ratio test.")
  }

  p_value <- as.numeric(body[nrow(body), p_column[1L]])
  if (!is.finite(p_value)) stop("The VGAM LRT returned an invalid p-value.")
  p_value
}

select_ppom_structure <- function(data,
                                  formula = RET ~ SM + DIAB + GH + BP,
                                  initial_effects = c("SM", "DIAB"),
                                  alpha = 0.05) {
  current_formula <- formula
  current_xvars <- attr(stats::terms(formula), "term.labels")
  current_effects <- initial_effects

  current_pom <- fit_ppom_model(
    current_formula, data, nonparallel = character(0)
  )
  current_fit <- fit_ppom_model(
    current_formula, data, nonparallel = current_effects,
    etastart = current_pom@predictors
  )

  history <- data.frame(
    component = character(), removed = character(), p_value = numeric()
  )

  repeat {
    removed_in_pass <- FALSE

    # First test category-specific effects while retaining all beta terms.
    repeat {
      if (length(current_effects) == 0L) break

      reduced_fits <- setNames(vector("list", length(current_effects)),
                               current_effects)
      p_values <- setNames(numeric(length(current_effects)), current_effects)

      for (variable in current_effects) {
        reduced_fits[[variable]] <- fit_ppom_model(
          current_formula,
          data,
          nonparallel = setdiff(current_effects, variable),
          etastart = current_pom@predictors
        )
        p_values[variable] <- extract_vgam_lrt_p(
          VGAM::lrtest(reduced_fits[[variable]], current_fit)
        )
      }

      variable <- names(which.max(p_values))
      p_value <- unname(p_values[variable])
      if (p_value <= alpha) break

      history <- rbind(history, data.frame(
        component = "gamma", removed = variable, p_value = p_value
      ))
      current_effects <- setdiff(current_effects, variable)
      current_fit <- reduced_fits[[variable]]
      removed_in_pass <- TRUE
    }

    # A beta term is eligible for removal only after its gamma is absent.
    repeat {
      beta_candidates <- setdiff(current_xvars, current_effects)
      if (length(beta_candidates) == 0L) break

      reduced_fits <- setNames(vector("list", length(beta_candidates)),
                               beta_candidates)
      reduced_formulas <- setNames(vector("list", length(beta_candidates)),
                                   beta_candidates)
      reduced_pom <- setNames(vector("list", length(beta_candidates)),
                              beta_candidates)
      p_values <- setNames(numeric(length(beta_candidates)), beta_candidates)

      for (variable in beta_candidates) {
        reduced_formulas[[variable]] <- stats::update.formula(
          current_formula, paste(". ~ . -", variable)
        )
        reduced_pom[[variable]] <- fit_ppom_model(
          reduced_formulas[[variable]], data, nonparallel = character(0)
        )
        reduced_fits[[variable]] <- fit_ppom_model(
          reduced_formulas[[variable]],
          data,
          nonparallel = current_effects,
          etastart = reduced_pom[[variable]]@predictors
        )
        p_values[variable] <- extract_vgam_lrt_p(
          VGAM::lrtest(reduced_fits[[variable]], current_fit)
        )
      }

      variable <- names(which.max(p_values))
      p_value <- unname(p_values[variable])
      if (p_value <= alpha) break

      history <- rbind(history, data.frame(
        component = "beta", removed = variable, p_value = p_value
      ))
      current_xvars <- setdiff(current_xvars, variable)
      current_formula <- reduced_formulas[[variable]]
      current_pom <- reduced_pom[[variable]]
      current_fit <- reduced_fits[[variable]]
      removed_in_pass <- TRUE
    }

    if (!removed_in_pass) break
  }

  list(
    model = current_fit,
    formula = current_formula,
    xvars = current_xvars,
    effects = current_effects,
    history = history
  )
}
```

## Location-shift utilities

```{r location-shift-utilities}
select_location_shift_structure <- function(
    data,
    response = "RET",
    initial_xvars = c("SM", "DIAB", "GH", "BP"),
    initial_zvars = c("SM", "DIAB"),
    alpha = 0.05) {
  xvars <- initial_xvars
  zvars <- initial_zvars

  make_formula <- function() {
    x_part <- paste(xvars, collapse = " + ")
    if (length(zvars) == 0L) {
      stats::as.formula(paste(response, "~", x_part))
    } else {
      stats::as.formula(paste0(
        response, " ~ ", x_part, " | ", paste(zvars, collapse = " + ")
      ))
    }
  }

  fit_model <- function(etastart = NULL) {
    arguments <- list(
      formula = make_formula(),
      data = data,
      family = "cumulative",
      scaling = TRUE,
      middle = TRUE,
      reverse = FALSE,
      trace = FALSE
    )
    if (!is.null(etastart)) arguments$etastart <- etastart
    do.call(ordDisp::ordDisp, arguments)
  }

  location_only_formula <- stats::as.formula(
    paste(response, "~", paste(xvars, collapse = " + "))
  )
  location_only_fit <- ordDisp::ordDisp(
    location_only_formula,
    data = data,
    family = "cumulative",
    scaling = TRUE,
    middle = TRUE,
    reverse = FALSE,
    trace = FALSE
  )
  fit <- fit_model(etastart = location_only_fit@predictors)

  history <- data.frame(
    component = character(), removed = character(), p_value = numeric()
  )

  repeat {
    removed_in_pass <- FALSE

    # Shift effects are selected first; location effects are then selected
    # regardless of whether the corresponding shift effect remains.
    for (side in c("z", "x")) {
      repeat {
        current_vars <- if (side == "z") zvars else xvars
        if (length(current_vars) == 0L) break
        if (side == "x" && length(xvars) == 1L) break

        drop_table <- stats::drop1(fit, test = "LRT")
        p_column <- grep("^Pr\\(", names(drop_table), value = TRUE)
        if (length(p_column) == 0L) {
          stop("No LRT p-value column was found in drop1(ordDisp fit).")
        }

        candidate_terms <- paste0(current_vars, side)
        index <- which(
          !is.na(drop_table[[p_column[1L]]]) &
            rownames(drop_table) %in% candidate_terms
        )
        if (length(index) == 0L) break

        index <- index[which.max(drop_table[[p_column[1L]]][index])]
        term <- rownames(drop_table)[index]
        p_value <- drop_table[[p_column[1L]]][index]
        if (!is.finite(p_value)) stop("Invalid ordDisp LRT p-value.")
        if (p_value <= alpha) break

        variable <- sub("[xz]$", "", term)
        history <- rbind(history, data.frame(
          component = if (side == "z") "gamma" else "beta",
          removed = variable,
          p_value = p_value
        ))

        if (side == "z") {
          zvars <- setdiff(zvars, variable)
        } else {
          xvars <- setdiff(xvars, variable)
        }

        fit <- fit_model(etastart = fit@predictors)
        removed_in_pass <- TRUE
      }
    }

    if (!removed_in_pass) break
  }

  list(
    model = fit,
    xvars = xvars,
    zvars = zvars,
    history = history
  )
}

predict_location_shift <- function(fit, newdata, xvars, zvars,
                                   outcome_levels = c("0", "1", "2")) {
  coefficients <- stats::coef(fit)
  required <- c(
    "(Intercept):1", "(Intercept):2",
    paste0(xvars, "x"),
    if (length(zvars) > 0L) paste0(zvars, "z") else character(0)
  )
  missing <- setdiff(required, names(coefficients))
  if (length(missing) > 0L) {
    stop("Missing ordDisp coefficients: ", paste(missing, collapse = ", "))
  }

  X <- as.matrix(newdata[, xvars, drop = FALSE])
  location <- drop(X %*% unname(coefficients[paste0(xvars, "x")]))

  if (length(zvars) > 0L) {
    Z <- as.matrix(newdata[, zvars, drop = FALSE])
    shift <- drop(Z %*% unname(coefficients[paste0(zvars, "z")]))
  } else {
    shift <- rep(0, nrow(newdata))
  }

  eta_1 <- unname(coefficients["(Intercept):1"]) + location - 0.5 * shift
  eta_2 <- unname(coefficients["(Intercept):2"]) + location + 0.5 * shift
  cumulative_1 <- stats::plogis(eta_1)
  cumulative_2 <- stats::plogis(eta_2)

  prepare_prob(
    cbind(cumulative_1, cumulative_2 - cumulative_1, 1 - cumulative_2),
    outcome_levels
  )
}
```

## GPPOM utilities

```{r gppom-utilities}
is_identifiable_gppom <- function(surv_name, rhoj_name) {
  !(surv_name == "H4" &&
      rhoj_name %in% c("rho_14", "rho_44", "rho_54"))
}

make_category_matrix <- function(data, effect_vars) {
  if (length(effect_vars) == 0L) return(NULL)

  out <- as.matrix(data[, effect_vars, drop = FALSE])
  storage.mode(out) <- "double"
  colnames(out) <- effect_vars
  out
}

is_usable_gppom_fit <- function(fit, criterion = NULL) {
  if (inherits(fit, "error") || is.null(fit)) return(FALSE)

  usable <- length(fit$conv) == 1L && fit$conv == 0L &&
    length(fit$nll) == 1L && is.finite(fit$nll) && fit$nll < 1e11

  if (!is.null(criterion)) {
    field <- tolower(criterion)
    usable <- usable && length(fit[[field]]) == 1L &&
      is.finite(fit[[field]])
  }

  isTRUE(usable)
}

fit_gppom_structure <- function(X, y, m, data, effect_vars,
                                 surv_name, rhoj_name,
                                 surv_candidates, rhoj_candidates) {
  eta_name <- get_eta_name_for_rhoj(rhoj_name)
  surv_info <- surv_candidates[[surv_name]]
  eta_info <- geta_candidates[[eta_name]]
  rhoj_info <- rhoj_candidates[[rhoj_name]]

  initial <- get_data_driven_init(
    y = y,
    m = m,
    surv_info = surv_info,
    eta_info = eta_info,
    rhoj_info = rhoj_info
  )
  surv_info$init <- initial$surv_init
  rhoj_info$init <- initial$rhoj_init

  fit <- fit_one_combo(
    X = X,
    y = y,
    m = m,
    n = nrow(X),
    surv_info = surv_info,
    eta_info = eta_info,
    rhoj_info = rhoj_info,
    T_np = make_category_matrix(data, effect_vars),
    method = "BFGS"
  )

  c(list(
    surv = surv_name,
    geta = eta_name,
    rhoj = rhoj_name,
    effect_vars = effect_vars
  ), fit)
}

fit_all_gppom_candidates <- function(
    X, y, m, data, effect_vars,
    surv_names = names(make_surv_func(m))) {
  surv_candidates <- make_surv_func(m)
  rhoj_candidates <- make_rhoj_candidates()
  all_results <- list()

  for (surv_name in surv_names) {
    for (rhoj_name in names(rhoj_candidates)) {
      if (!is_identifiable_gppom(surv_name, rhoj_name)) next

      eta_name <- get_eta_name_for_rhoj(rhoj_name)
      key <- paste(surv_name, eta_name, rhoj_name, sep = "__")

      all_results[[key]] <- tryCatch(
        fit_gppom_structure(
          X = X,
          y = y,
          m = m,
          data = data,
          effect_vars = effect_vars,
          surv_name = surv_name,
          rhoj_name = rhoj_name,
          surv_candidates = surv_candidates,
          rhoj_candidates = rhoj_candidates
        ),
        error = function(error) error
      )
    }
  }

  list(
    all_results = all_results,
    surv_candidates = surv_candidates,
    rhoj_candidates = rhoj_candidates
  )
}

select_gppom_candidate <- function(bundle, criterion = c("AIC", "BIC")) {
  criterion <- match.arg(criterion)
  successful <- bundle$all_results[
    vapply(
      bundle$all_results,
      is_usable_gppom_fit,
      logical(1),
      criterion = criterion
    )
  ]

  if (length(successful) == 0L) {
    stop("No GPPOM candidate converged with a finite ", criterion, ".")
  }

  # select_best_by_rhoa() first chooses the best rho_l/baseline within
  # each rho_a family. The final line then chooses one overall model.
  best_by_rhoa <- select_best_by_rhoa(
    all_results = successful,
    criterion = criterion,
    require_convergence = TRUE
  )
  selected_row <- best_by_rhoa[
    which.min(best_by_rhoa[[criterion]]), , drop = FALSE
  ]

  list(
    key = selected_row$key[[1L]],
    row = selected_row,
    fit = successful[[selected_row$key[[1L]]]],
    criterion = criterion
  )
}

backward_gppom <- function(selected, X, y, m, data,
                            initial_effects = c("SM", "DIAB"),
                            surv_candidates, rhoj_candidates,
                            alpha = 0.05,
                            numerical_tolerance = 1e-6) {
  current_X <- X
  current_xvars <- colnames(X)
  current_effects <- initial_effects
  current_fit <- selected$fit
  surv_name <- current_fit$surv
  rhoj_name <- current_fit$rhoj

  history <- data.frame(
    component = character(), removed = character(),
    statistic = numeric(), df = integer(), p_value = numeric()
  )

  evaluate_reductions <- function(candidates, component) {
    fits <- setNames(vector("list", length(candidates)), candidates)
    statistic <- setNames(numeric(length(candidates)), candidates)
    degrees <- setNames(integer(length(candidates)), candidates)
    p_value <- setNames(numeric(length(candidates)), candidates)

    for (variable in candidates) {
      reduced_effects <- if (component == "gamma") {
        setdiff(current_effects, variable)
      } else {
        current_effects
      }
      reduced_xvars <- if (component == "beta") {
        setdiff(current_xvars, variable)
      } else {
        current_xvars
      }

      fit <- fit_gppom_structure(
        X = current_X[, reduced_xvars, drop = FALSE],
        y = y,
        m = m,
        data = data,
        effect_vars = reduced_effects,
        surv_name = surv_name,
        rhoj_name = rhoj_name,
        surv_candidates = surv_candidates,
        rhoj_candidates = rhoj_candidates
      )

      if (!is_usable_gppom_fit(fit)) {
        stop("A reduced GPPOM did not converge after removing ",
             component, " for ", variable, ".")
      }

      value <- 2 * (fit$nll - current_fit$nll)
      df_difference <- length(current_fit$par) - length(fit$par)

      if (df_difference <= 0L) stop("Non-positive GPPOM LRT degrees of freedom.")
      if (value < -numerical_tolerance) {
        stop("A reduced GPPOM has a materially larger likelihood.")
      }

      value <- max(0, value)
      fits[[variable]] <- fit
      statistic[variable] <- value
      degrees[variable] <- df_difference
      p_value[variable] <- stats::pchisq(
        value, df = df_difference, lower.tail = FALSE
      )
    }

    list(fits = fits, statistic = statistic, df = degrees, p_value = p_value)
  }

  repeat {
    removed_in_pass <- FALSE

    # First remove category-specific effects while retaining all beta terms.
    repeat {
      if (length(current_effects) == 0L) break
      reductions <- evaluate_reductions(current_effects, "gamma")
      variable <- names(which.max(reductions$p_value))
      p_value <- unname(reductions$p_value[variable])
      if (p_value <= alpha) break

      history <- rbind(history, data.frame(
        component = "gamma",
        removed = variable,
        statistic = unname(reductions$statistic[variable]),
        df = unname(reductions$df[variable]),
        p_value = p_value
      ))
      current_effects <- setdiff(current_effects, variable)
      current_fit <- reductions$fits[[variable]]
      removed_in_pass <- TRUE
    }

    # A beta term is eligible only after its gamma is absent.
    repeat {
      beta_candidates <- setdiff(current_xvars, current_effects)
      if (length(beta_candidates) == 0L) break
      reductions <- evaluate_reductions(beta_candidates, "beta")
      variable <- names(which.max(reductions$p_value))
      p_value <- unname(reductions$p_value[variable])
      if (p_value <= alpha) break

      history <- rbind(history, data.frame(
        component = "beta",
        removed = variable,
        statistic = unname(reductions$statistic[variable]),
        df = unname(reductions$df[variable]),
        p_value = p_value
      ))
      current_xvars <- setdiff(current_xvars, variable)
      current_X <- current_X[, current_xvars, drop = FALSE]
      current_fit <- reductions$fits[[variable]]
      removed_in_pass <- TRUE
    }

    if (!removed_in_pass) break
  }

  list(
    model = current_fit,
    xvars = current_xvars,
    effects = current_effects,
    history = history
  )
}

predict_gppom <- function(X, data, fit, effect_vars, m,
                           outcome_levels, surv_candidates,
                           rhoj_candidates) {
  eta_common <- drop(X %*% fit$beta)

  if (length(effect_vars) > 0L) {
    eta <- matrix(eta_common, nrow = nrow(X), ncol = m - 1L) +
      make_category_matrix(data, effect_vars) %*% fit$Gamma
  } else {
    eta <- eta_common
  }

  eta_value <- geta_candidates[[fit$geta]]$func(eta)
  if (is.null(dim(eta_value))) {
    eta_value <- matrix(eta_value, nrow = nrow(X), ncol = m - 1L)
  }

  Q <- vapply(seq_len(m - 1L), function(j) {
    rho <- do.call(
      rhoj_candidates[[fit$rhoj]]$fun,
      c(
        list(eta_val = eta_value[, j], j = j, m = m),
        as.list(fit$rhoj_par)
      )
    )
    do.call(
      surv_candidates[[fit$surv]]$func,
      c(list(t = rho), as.list(fit$surv_par))
    )
  }, FUN.VALUE = numeric(nrow(X)))

  Q <- matrix(Q, nrow = nrow(X), ncol = m - 1L)
  prepare_prob(cbind(1, Q) - cbind(Q, 0), outcome_levels)
}
```

## Complete-data fits for the upper part of Table 6

```{r complete-data-fits}
retino_formula <- RET ~ SM + DIAB + GH + BP

# PPOM: SM and DIAB have category-specific effects.
full_pom_start <- fit_ppom_model(
  retino_formula, retino_full, nonparallel = character(0)
)
full_ppom <- fit_ppom_model(
  retino_formula,
  retino_full,
  nonparallel = c("SM", "DIAB"),
  etastart = full_pom_start@predictors
)

# Location-shift model used in Table 6. SM has a shift effect but no
# separate location coefficient; DIAB has both location and shift effects.
full_location_shift <- ordDisp::ordDisp(
  RET ~ DIAB + GH + BP | SM + DIAB,
  data = retino_full,
  family = "cumulative",
  scaling = TRUE,
  middle = TRUE,
  reverse = FALSE,
  trace = FALSE
)

# The complete-data GPPOMs in Table 6 use H1 and a category-specific
# effect for SM only. This gives 7 parameters for the AIC model and
# 6 parameters for the BIC model, as reported in the paper.
X_full <- stats::model.matrix(
  ~ -1 + SM + DIAB + GH + BP,
  data = retino_full
)
y_full <- as.integer(retino_full$RET)
m <- nlevels(retino_full$RET)

full_gppom_bundle <- fit_all_gppom_candidates(
  X = X_full,
  y = y_full,
  m = m,
  data = retino_full,
  effect_vars = "SM",
  surv_names = "H1"
)

full_gppom_AIC <- select_gppom_candidate(full_gppom_bundle, "AIC")
full_gppom_BIC <- select_gppom_candidate(full_gppom_bundle, "BIC")

gppom_selections <- rbind(
  data.frame(Criterion = "AIC", full_gppom_AIC$row),
  data.frame(Criterion = "BIC", full_gppom_BIC$row)
)

print(gppom_selections[, c(
  "Criterion", "surv", "rhoj", "logLik", "AIC", "BIC"
)], row.names = FALSE)
```

## Extract the complete-data estimates

```{r extract-complete-data-results}
extract_ppom_result <- function(fit) {
  estimate <- stats::coef(fit)
  covariance <- stats::vcov(fit)

  transformed_effect <- function(variable) {
    name_1 <- paste0(variable, ":1")
    name_2 <- paste0(variable, ":2")

    # With reverse = FALSE, VGAM reports the slopes for P(Y <= j).
    # Table 6 uses P(Y > j), so its beta is -b1. Because b2 represents
    # -(beta + gamma), the reported gamma is b1 - b2.
    beta <- -unname(estimate[name_1])
    gamma <- unname(estimate[name_1] - estimate[name_2])
    se_beta <- sqrt(covariance[name_1, name_1])
    se_gamma <- sqrt(
      covariance[name_1, name_1] + covariance[name_2, name_2] -
        2 * covariance[name_1, name_2]
    )

    c(beta = beta, se_beta = se_beta, gamma = gamma, se_gamma = se_gamma)
  }

  parallel_effect <- function(variable) {
    c(
      beta = -unname(estimate[variable]),
      se_beta = sqrt(covariance[variable, variable])
    )
  }

  sm <- transformed_effect("SM")
  diab <- transformed_effect("DIAB")
  gh <- parallel_effect("GH")
  bp <- parallel_effect("BP")
  log_likelihood <- as.numeric(stats::logLik(fit))

  list(
    values = c(
      format_estimate(sm["beta"], sm["se_beta"]),
      format_estimate(sm["gamma"], sm["se_gamma"]),
      format_estimate(diab["beta"], diab["se_beta"]),
      format_estimate(diab["gamma"], diab["se_gamma"]),
      format_estimate(gh["beta"], gh["se_beta"]),
      format_estimate(bp["beta"], bp["se_beta"]),
      as.character(model_df(fit)),
      format_number(log_likelihood),
      format_number(AIC(fit)),
      format_number(BIC(fit))
    )
  )
}

extract_location_shift_result <- function(fit) {
  estimate <- stats::coef(fit)
  covariance <- stats::vcov(fit)

  get_effect <- function(coefficient_name) {
    c(
      estimate = -unname(estimate[coefficient_name]),
      se = sqrt(covariance[coefficient_name, coefficient_name])
    )
  }

  sm_gamma <- get_effect("SMz")
  diab_beta <- get_effect("DIABx")
  diab_gamma <- get_effect("DIABz")
  gh_beta <- get_effect("GHx")
  bp_beta <- get_effect("BPx")
  log_likelihood <- as.numeric(stats::logLik(fit))

  list(
    values = c(
      "-",
      format_estimate(sm_gamma["estimate"], sm_gamma["se"]),
      format_estimate(diab_beta["estimate"], diab_beta["se"]),
      format_estimate(diab_gamma["estimate"], diab_gamma["se"]),
      format_estimate(gh_beta["estimate"], gh_beta["se"]),
      format_estimate(bp_beta["estimate"], bp_beta["se"]),
      as.character(model_df(fit)),
      format_number(log_likelihood),
      format_number(AIC(fit)),
      format_number(BIC(fit))
    )
  )
}

extract_gppom_result <- function(selected, X, effect_vars) {
  fit <- selected$fit
  names(fit$beta) <- colnames(X)
  names(fit$se_beta) <- colnames(X)

  gamma <- setNames(rep(NA_real_, length(effect_vars)), effect_vars)
  gamma_se <- gamma
  if (length(effect_vars) > 0L) {
    gamma[] <- fit$gamma_free
    gamma_se[] <- fit$se_gamma
  }

  value_or_dash <- function(variable, component = c("beta", "gamma")) {
    component <- match.arg(component)

    if (component == "beta") {
      if (!(variable %in% names(fit$beta))) return("-")
      return(format_estimate(fit$beta[variable], fit$se_beta[variable]))
    }

    if (!(variable %in% names(gamma))) return("-")
    format_estimate(gamma[variable], gamma_se[variable])
  }

  list(
    values = c(
      value_or_dash("SM", "beta"),
      value_or_dash("SM", "gamma"),
      value_or_dash("DIAB", "beta"),
      value_or_dash("DIAB", "gamma"),
      value_or_dash("GH", "beta"),
      value_or_dash("BP", "beta"),
      as.character(length(fit$par)),
      format_number(-fit$nll),
      format_number(fit$aic),
      format_number(fit$bic)
    )
  )
}

ppom_complete <- extract_ppom_result(full_ppom)$values
location_complete <- extract_location_shift_result(full_location_shift)$values
gppom_AIC_complete <- extract_gppom_result(
  full_gppom_AIC, X_full, effect_vars = "SM"
)$values
gppom_BIC_complete <- extract_gppom_result(
  full_gppom_BIC, X_full, effect_vars = "SM"
)$values
```

## Fold-specific fits and predictions

```{r fold-specific-fits}
fit_ppom_fold <- function(train, test, outcome_levels, alpha = 0.05) {
  selected <- select_ppom_structure(
    data = train,
    initial_effects = c("SM", "DIAB"),
    alpha = alpha
  )

  list(
    prob = prepare_prob(
      VGAM::predictvglm(selected$model, newdata = test, type = "response"),
      outcome_levels
    ),
    selected = selected
  )
}

fit_location_shift_fold <- function(train, test, outcome_levels,
                                    alpha = 0.05) {
  selected <- select_location_shift_structure(
    data = train,
    initial_xvars = c("SM", "DIAB", "GH", "BP"),
    initial_zvars = c("SM", "DIAB"),
    alpha = alpha
  )

  list(
    prob = predict_location_shift(
      selected$model,
      test,
      selected$xvars,
      selected$zvars,
      outcome_levels
    ),
    selected = selected
  )
}

fit_gppom_BIC_fold <- function(train, test, outcome_levels,
                               alpha = 0.05) {
  m <- length(outcome_levels)
  y_train <- match(as.character(train$RET), outcome_levels)

  X_train <- stats::model.matrix(
    ~ -1 + SM + DIAB + GH + BP,
    data = train
  )
  X_test <- stats::model.matrix(
    ~ -1 + SM + DIAB + GH + BP,
    data = test
  )
  X_test <- X_test[, colnames(X_train), drop = FALSE]

  maximal_bundle <- fit_all_gppom_candidates(
    X = X_train,
    y = y_train,
    m = m,
    data = train,
    effect_vars = c("SM")
  )
  maximal_selected <- select_gppom_candidate(maximal_bundle, "BIC")

  final_selected <- backward_gppom(
    selected = maximal_selected,
    X = X_train,
    y = y_train,
    m = m,
    data = train,
    initial_effects = c("SM"),
    surv_candidates = maximal_bundle$surv_candidates,
    rhoj_candidates = maximal_bundle$rhoj_candidates,
    alpha = alpha
  )

  probability <- predict_gppom(
    X = X_test[, final_selected$xvars, drop = FALSE],
    data = test,
    fit = final_selected$model,
    effect_vars = final_selected$effects,
    m = m,
    outcome_levels = outcome_levels,
    surv_candidates = maximal_bundle$surv_candidates,
    rhoj_candidates = maximal_bundle$rhoj_candidates
  )

  list(
    prob = probability,
    selected_candidate = maximal_selected$key,
    selected_xvars = final_selected$xvars,
    selected_effects = final_selected$effects,
    history = final_selected$history
  )
}
```

## Repeated stratified cross-validation

```{r cross-validation, warning=FALSE}
run_retinopathy_cv <- function(
    data,
    folds,
    continuous_vars = c("DIAB", "GH", "BP"),
    alpha = 0.05) {
  outcome_levels <- levels(data$RET)
  predictions <- list()
  failures <- list()
  counter <- 0L
  failure_counter <- 0L

  for (repeat_id in sort(unique(folds$repeat_id))) {
    for (fold_id in sort(unique(folds$fold_id))) {
      cat(sprintf(
        "[CV] repeat %d, fold %d\n", repeat_id, fold_id
      ))

      test_ids <- folds$row_id[
        folds$repeat_id == repeat_id & folds$fold_id == fold_id
      ]
      train_raw <- data[!data$.row_id %in% test_ids, , drop = FALSE]
      test_raw <- data[data$.row_id %in% test_ids, , drop = FALSE]

      preprocessor <- fit_preprocessor(train_raw, continuous_vars)
      train <- apply_preprocessor(train_raw, preprocessor)
      test <- apply_preprocessor(test_raw, preprocessor)

      fitters <- list(
        PPOM = fit_ppom_fold,
        `Location-shift` = fit_location_shift_fold,
        GPPOM_BIC = fit_gppom_BIC_fold
      )

      for (model_name in names(fitters)) {
        result <- tryCatch(
          fitters[[model_name]](
            train = train,
            test = test,
            outcome_levels = outcome_levels,
            alpha = alpha
          ),
          error = function(error) error
        )

        if (inherits(result, "error")) {
          failure_counter <- failure_counter + 1L
          failures[[failure_counter]] <- data.frame(
            repeat_id = repeat_id,
            fold_id = fold_id,
            model = model_name,
            message = conditionMessage(result)
          )
          next
        }

        probability <- prepare_prob(
          result$prob,
          outcome_levels
        )
        scores <- observation_scores(
          test$RET,
          probability,
          outcome_levels
        )

        counter <- counter + 1L
        predictions[[counter]] <- data.frame(
          repeat_id = repeat_id,
          fold_id = fold_id,
          row_id = test$.row_id,
          model = model_name,
          scores
        )
      }
    }
  }

  prediction_data <- if (length(predictions) > 0L) {
    do.call(rbind, predictions)
  } else {
    data.frame()
  }

  failure_data <- if (length(failures) > 0L) {
    do.call(rbind, failures)
  } else {
    data.frame(
      repeat_id = integer(), fold_id = integer(),
      model = character(), message = character()
    )
  }

  list(predictions = prediction_data, failures = failure_data)
}

summarise_cv_scores <- function(cv_result) {
  if (nrow(cv_result$failures) > 0L) {
    print(cv_result$failures)
    stop("At least one model-fold fit failed; scores were not summarized.")
  }

  by_repeat <- stats::aggregate(
    cbind(LogS, RPS) ~ model + repeat_id,
    data = cv_result$predictions,
    FUN = mean
  )

  overall <- stats::aggregate(
    cbind(LogS, RPS) ~ model,
    data = by_repeat,
    FUN = mean
  )
  overall[match(c("PPOM", "Location-shift", "GPPOM_BIC"), overall$model), ]
}

folds <- make_stratified_folds(
  y = retino_raw$RET,
  v = 10L,
  repeats = 10L,
  seed = 1234L
)

cv_result <- run_retinopathy_cv(
  data = retino_raw,
  folds = folds,
  continuous_vars = continuous_vars,
  alpha = 0.05
)

cv_scores <- summarise_cv_scores(cv_result)
print(cv_scores, row.names = FALSE)
```

## Table 6 values in one console object

```{r table-6-console-output}
score_value <- function(model, score) {
  value <- cv_scores[cv_scores$model == model, score]
  if (length(value) != 1L) return(NA_character_)
  format_number(value, digits = 4L)
}

row_labels <- c(
  "SM (beta)", "SM (gamma)",
  "DIAB (beta)", "DIAB (gamma)",
  "GH (beta)", "BP (beta)",
  "# of param.", "logLik", "AIC", "BIC", "LogS", "RPS"
)

table6_results <- data.frame(
  Param. = row_labels,
  PPOM = c(
    ppom_complete,
    score_value("PPOM", "LogS"),
    score_value("PPOM", "RPS")
  ),
  `Location-shift` = c(
    location_complete,
    score_value("Location-shift", "LogS"),
    score_value("Location-shift", "RPS")
  ),
  GPPOM_AIC = c(
    gppom_AIC_complete,
    "-",
    "-"
  ),
  GPPOM_BIC = c(
    gppom_BIC_complete,
    score_value("GPPOM_BIC", "LogS"),
    score_value("GPPOM_BIC", "RPS")
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)

names(table6_results)[4:5] <- c(
  paste0(
    "GPPOM AIC (", full_gppom_AIC$row$surv[[1L]], ", ",
    full_gppom_AIC$row$rhoj[[1L]], ")"
  ),
  paste0(
    "GPPOM BIC (", full_gppom_BIC$row$surv[[1L]], ", ",
    full_gppom_BIC$row$rhoj[[1L]], ")"
  )
)

print(table6_results, row.names = FALSE, right = FALSE)
```

