---
title: "Supplementary Simulation Study for the GPPOM"
output:
  html_document:
    code_folding: show
    highlight: textmate
    df_print: paged
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

# Common code

```{r setup, message=FALSE, warning=FALSE}
library(VGAM)
library(zipfR)
library(dplyr)
library(ggplot2)
library(patchwork)

# This R Markdown file and common_functions.R should be in the same directory.
source(file.path("code", "common_functions.R"))

n_obs <- 500L
n_rep <- 500L
m_cat <- 4L
base_seed <- 10000L
```

## Probability and data-generation functions

```{r data-generation-functions}
# Validate and standardize the ordering of category-probability columns.
prepare_prob <- function(prob, outcome_levels, tolerance = 1e-8) {
  prob <- as.matrix(prob)
  outcome_levels <- as.character(outcome_levels)

  if (ncol(prob) != length(outcome_levels)) {
    stop("The number of probability columns does not match the outcome levels.")
  }

  if (!is.null(colnames(prob)) &&
      all(outcome_levels %in% colnames(prob))) {
    prob <- prob[, outcome_levels, drop = FALSE]
  } else {
    colnames(prob) <- outcome_levels
  }

  if (any(!is.finite(prob))) {
    stop("The predicted probabilities contain NA, NaN, or Inf values.")
  }

  if (any(prob < -tolerance) || any(prob > 1 + tolerance)) {
    stop("The predicted probabilities fall outside [0, 1].")
  }

  prob[prob < 0] <- 0
  prob[prob > 1] <- 1

  row_totals <- rowSums(prob)

  if (any(abs(row_totals - 1) > tolerance)) {
    stop("The rows of the predicted probability matrix do not sum to one.")
  }

  prob / row_totals
}


# True conditional survival probabilities under the PPOM.
true_Q_ppom <- function(dat,
                        alpha = c(-1.5, 0, 1.5),
                        beta = c(0.7, 0.9),
                        gamma = c(0, 0.2, -1)) {
  n <- nrow(dat)
  m <- length(alpha) + 1L

  eta_common <- beta[1] * dat$x1 + beta[2] * dat$x2
  eta_matrix <- matrix(
    eta_common,
    nrow = n,
    ncol = m - 1L
  ) + dat$x2 %o% gamma

  cumulative_prob <- plogis(
    matrix(alpha, nrow = n, ncol = m - 1L, byrow = TRUE) -
      eta_matrix
  )

  Q <- 1 - cumulative_prob
  colnames(Q) <- paste0("Q", seq_len(m - 1L))
  Q
}


# True conditional survival probabilities under the location-shift model.
true_Q_location_shift <- function(dat,
                                  alpha = c(-1.5, 0, 1.5),
                                  beta = c(0.7, 0.4),
                                  delta = 0.5) {
  n <- nrow(dat)
  m <- length(alpha) + 1L
  shift_scale <- seq_len(m - 1L) - m / 2

  cumulative_predictor <-
    matrix(alpha, nrow = n, ncol = m - 1L, byrow = TRUE) -
    beta[1] * dat$x1 -
    beta[2] * dat$x2 +
    dat$x2 %o% (shift_scale * delta)

  Q <- 1 - plogis(cumulative_predictor)
  colnames(Q) <- paste0("Q", seq_len(m - 1L))
  Q
}


# True conditional survival probabilities under S2{rho_1,1(x; j)}.
true_Q_gppom <- function(dat,
                         m = 4L,
                         beta = c(0.7, 0.9),
                         gamma = c(0, 0.2, -1)) {
  n <- nrow(dat)

  if (length(gamma) != m - 1L) {
    stop("The length of gamma must equal m - 1.")
  }

  eta_common <- beta[1] * dat$x1 + beta[2] * dat$x2
  eta_matrix <- matrix(
    eta_common,
    nrow = n,
    ncol = m - 1L
  ) + dat$x2 %o% gamma

  Q <- vapply(
    seq_len(m - 1L),
    function(j) {
      (1 - (j / m)^exp(eta_matrix[, j]))^2
    },
    FUN.VALUE = numeric(n)
  )

  Q <- matrix(Q, nrow = n, ncol = m - 1L)
  colnames(Q) <- paste0("Q", seq_len(m - 1L))
  Q
}


# Generate one ordinal data set from supplied conditional survival probabilities.
generate_ordinal_data <- function(n, m, true_Q_function, seed) {
  set.seed(seed)

  dat <- data.frame(
    x1 = runif(n, min = -1, max = 1),
    x2 = rbinom(n, size = 1, prob = 0.5)
  )

  Q_true <- as.matrix(true_Q_function(dat))

  if (!all(dim(Q_true) == c(n, m - 1L))) {
    stop("The true survival-probability matrix has an invalid dimension.")
  }

  if (any(!is.finite(Q_true)) ||
      any(Q_true <= 0) ||
      any(Q_true >= 1) ||
      any(Q_true[, seq_len(m - 2L), drop = FALSE] <=
            Q_true[, 2:(m - 1L), drop = FALSE])) {
    stop("The true conditional survival probabilities are invalid.")
  }

  cumulative_prob <- 1 - Q_true
  u <- runif(n)
  y <- 1L + rowSums(
    matrix(u, nrow = n, ncol = m - 1L) > cumulative_prob
  )

  dat$y <- factor(
    y,
    levels = seq_len(m),
    ordered = TRUE
  )

  list(data = dat, Q_true = Q_true)
}
```

## Prediction and evaluation functions

```{r prediction-functions}
# Convert category probabilities to conditional survival probabilities.
prob_to_Q <- function(prob) {
  prob <- as.matrix(prob)
  m <- ncol(prob)

  Q <- sapply(seq_len(m - 1L), function(j) {
    rowSums(prob[, (j + 1L):m, drop = FALSE])
  })

  Q <- as.matrix(Q)
  colnames(Q) <- paste0("Q", seq_len(m - 1L))
  Q
}


# Mean squared error of the fitted conditional survival probabilities.
calc_MSE_Q <- function(Q_hat, Q_true) {
  mean((Q_hat - Q_true)^2)
}


# Fitted conditional survival probabilities from the PPOM.
get_Q_ppom <- function(model, dat) {
  prob <- VGAM::predictvglm(
    model,
    newdata = dat[, c("x1", "x2"), drop = FALSE],
    type = "response"
  )

  prob <- prepare_prob(
    prob,
    outcome_levels = levels(dat$y)
  )

  prob_to_Q(prob)
}


# Fitted conditional survival probabilities from the location-shift model.
get_Q_location_shift <- function(model, dat) {
  prob <- prepare_prob(
    fitted(model),
    outcome_levels = levels(dat$y)
  )

  prob_to_Q(prob)
}


# Fitted conditional survival probabilities from the GPPOM.
get_Q_gppom <- function(fit, X, T_np, m,
                        outcome_levels,
                        surv_candidates,
                        rhoj_candidates) {
  n <- nrow(X)
  eta_common <- drop(X %*% fit$beta)

  eta_matrix <- matrix(
    eta_common,
    nrow = n,
    ncol = m - 1L
  ) + T_np %*% fit$Gamma

  eta_value <- geta_candidates[[fit$geta]]$func(eta_matrix)

  Q <- vapply(
    seq_len(m - 1L),
    function(j) {
      t_j <- do.call(
        rhoj_candidates[[fit$rhoj]]$fun,
        c(
          list(
            eta_val = eta_value[, j],
            j = j,
            m = m
          ),
          as.list(fit$rhoj_par)
        )
      )

      do.call(
        surv_candidates[[fit$surv]]$func,
        c(
          list(t = t_j),
          as.list(fit$surv_par)
        )
      )
    },
    FUN.VALUE = numeric(n)
  )

  Q <- matrix(Q, nrow = n, ncol = m - 1L)

  prob <- cbind(1, Q) - cbind(Q, 0)
  colnames(prob) <- outcome_levels
  prob <- prepare_prob(prob, outcome_levels)

  prob_to_Q(prob)
}
```

## Model-fitting functions

```{r model-fitting-functions}
# Fit the PPOM and location-shift competitors in one replication.
fit_competing_models <- function(dat, Q_true, rep_id) {
  fit_rows <- list()

  ppom_model <- VGAM::vglm(
    y ~ x1 + x2,
    family = VGAM::cumulative(
      link = "logitlink",
      parallel = FALSE ~ x2,
      reverse = FALSE
    ),
    data = dat
  )

  Q_hat <- get_Q_ppom(ppom_model, dat)

  fit_rows[["PPOM"]] <- data.frame(
    rep = rep_id,
    model = "PPOM",
    rho_a = NA_integer_,
    surv = NA_character_,
    rhoj = NA_character_,
    logLik = as.numeric(logLik(ppom_model)),
    AIC = AIC(ppom_model),
    BIC = BIC(ppom_model),
    MSE_Q = calc_MSE_Q(Q_hat, Q_true)
  )

  location_shift_model <- ordDisp::ordDisp(
    y ~ x1 + x2 | x2,
    data = dat,
    family = "cumulative",
    scaling = TRUE,
    middle = TRUE,
    reverse = FALSE
  )

  Q_hat <- get_Q_location_shift(location_shift_model, dat)

  fit_rows[["Location_shift"]] <- data.frame(
    rep = rep_id,
    model = "Location_shift",
    rho_a = NA_integer_,
    surv = NA_character_,
    rhoj = NA_character_,
    logLik = as.numeric(logLik(location_shift_model)),
    AIC = AIC(location_shift_model),
    BIC = BIC(location_shift_model),
    MSE_Q = calc_MSE_Q(Q_hat, Q_true)
  )

  dplyr::bind_rows(fit_rows)
}

# Exclude the nonidentifiable baseline-rho combinations.
is_identifiable_gppom <- function(surv_name, rhoj_name) {
  !(surv_name == "H4" &&
      rhoj_name %in% c("rho_14", "rho_44", "rho_54"))
}


# Construct the identifiable GPPOM candidate grid for one scenario.
make_gppom_grid <- function(scenario,
                            surv_candidates,
                            rhoj_candidates) {

  if (isTRUE(scenario$fit_all_gppoms)) {

    candidate_grid <- expand.grid(
      surv = names(surv_candidates),
      rhoj = names(rhoj_candidates),
      stringsAsFactors = FALSE
    )

    identifiable <- mapply(
      is_identifiable_gppom,
      candidate_grid$surv,
      candidate_grid$rhoj
    )

    candidate_grid <- candidate_grid[
      identifiable,
      ,
      drop = FALSE
    ]

    rownames(candidate_grid) <- NULL

    return(candidate_grid)
  }

  if (!is_identifiable_gppom(
    scenario$gppom_surv,
    scenario$gppom_rhoj
  )) {
    stop(
      "The fixed GPPOM specification is not identifiable: ",
      scenario$gppom_surv,
      " + ",
      scenario$gppom_rhoj
    )
  }

  data.frame(
    surv = scenario$gppom_surv,
    rhoj = scenario$gppom_rhoj,
    stringsAsFactors = FALSE
  )
}


# Fit the GPPOM candidates specified for one scenario.
fit_gppom_models <- function(dat, Q_true, m, rep_id, scenario) {
  n <- nrow(dat)
  X <- model.matrix(~ -1 + x1 + x2, data = dat)
  T_np <- model.matrix(~ -1 + x2, data = dat)
  y_num <- as.numeric(dat$y)

  surv_candidates <- make_surv_func(m = m)
  rhoj_candidates <- make_rhoj_candidates()
  candidate_grid <- make_gppom_grid(
    scenario = scenario,
    surv_candidates = surv_candidates,
    rhoj_candidates = rhoj_candidates
  )

  if (!isTRUE(scenario$fit_all_gppoms) &&
      nrow(candidate_grid) != 1L) {
    stop("A fixed GPPOM scenario must contain exactly one candidate.")
  }

  all_results <- list()

  for (i in seq_len(nrow(candidate_grid))) {
    surv_name <- candidate_grid$surv[i]
    rhoj_name <- candidate_grid$rhoj[i]
    eta_name <- get_eta_name_for_rhoj(rhoj_name)

    surv_info <- surv_candidates[[surv_name]]
    eta_info <- geta_candidates[[eta_name]]
    rhoj_info <- rhoj_candidates[[rhoj_name]]

    init_fit <- get_data_driven_init(
      y = y_num,
      m = m,
      surv_info = surv_info,
      eta_info = eta_info,
      rhoj_info = rhoj_info
    )

    surv_info$init <- init_fit$surv_init
    rhoj_info$init <- init_fit$rhoj_init

    fit <- fit_one_combo(
      X = X,
      y = y_num,
      m = m,
      n = n,
      surv_info = surv_info,
      eta_info = eta_info,
      rhoj_info = rhoj_info,
      T_np = T_np,
      method = "BFGS"
    )

    key <- paste(surv_name, eta_name, rhoj_name, sep = "__")

    all_results[[key]] <- c(
      list(
        surv = surv_name,
        geta = eta_name,
        rhoj = rhoj_name
      ),
      fit
    )
  }

  if (isTRUE(scenario$fit_all_gppoms)) {
    selected_by_rho <- select_best_by_rhoa(
      all_results = all_results,
      criterion = "BIC",
      require_convergence = TRUE
    )

    if (is.null(selected_by_rho) || nrow(selected_by_rho) == 0L) {
      stop("No converged GPPOM was available in replication ", rep_id, ".")
    }

    selected_fit <- selected_by_rho[
      which.min(selected_by_rho$BIC),
      ,
      drop = FALSE
    ]
    selected_key <- selected_fit$key[1]
  } else {
    selected_key <- names(all_results)[1]
    selected_model <- all_results[[selected_key]]

    if (is.null(selected_model$conv) || selected_model$conv != 0L) {
      stop("The fixed GPPOM did not converge in replication ", rep_id, ".")
    }
  }

  fit <- all_results[[selected_key]]
  rho_a <- parse_rhoj_name(fit$rhoj)$rho_a

  Q_hat <- get_Q_gppom(
    fit = fit,
    X = X,
    T_np = T_np,
    m = m,
    outcome_levels = levels(dat$y),
    surv_candidates = surv_candidates,
    rhoj_candidates = rhoj_candidates
  )

  data.frame(
    rep = rep_id,
    model = "GPPOM",
    rho_a = rho_a,
    surv = fit$surv,
    rhoj = fit$rhoj,
    logLik = -fit$nll,
    AIC = fit$aic,
    BIC = fit$bic,
    MSE_Q = calc_MSE_Q(Q_hat, Q_true)
  )
}


# Run one replication for a supplied scenario.
run_one_replication <- function(rep_id, scenario,
                                n = 500L, m = 4L,
                                seed = 10000L + rep_id) {
  simulated <- generate_ordinal_data(
    n = n,
    m = m,
    true_Q_function = scenario$true_Q_function,
    seed = seed
  )

  competitor_results <- fit_competing_models(
    dat = simulated$data,
    Q_true = simulated$Q_true,
    rep_id = rep_id
  )

  gppom_results <- fit_gppom_models(
    dat = simulated$data,
    Q_true = simulated$Q_true,
    m = m,
    rep_id = rep_id,
    scenario = scenario
  )

  dplyr::bind_rows(competitor_results, gppom_results)
}


# Run all replications for one scenario.
run_simulation <- function(scenario,
                           R = 500L,
                           n = 500L,
                           m = 4L,
                           seed_start = 10000L) {
  fit_list <- vector("list", R)
  failed_rep <- integer(0)

  for (r in seq_len(R)) {
    message(
      "Scenario ", scenario$id,
      ": replication ", r, " of ", R
    )

    result <- tryCatch(
      run_one_replication(
        rep_id = r,
        scenario = scenario,
        n = n,
        m = m,
        seed = seed_start + r
      ),
      error = function(e) {
        message(
          "Scenario ", scenario$id,
          ", replication ", r,
          " failed: ", conditionMessage(e)
        )
        NULL
      }
    )

    if (is.null(result)) {
      failed_rep <- c(failed_rep, r)
    } else {
      fit_list[[r]] <- result
    }
  }

  list(
    fit = dplyr::bind_rows(fit_list),
    failed_rep = failed_rep
  )
}
```

# Simulation scenarios

## Scenario (a): PPOM

Data are generated from the PPOM with
\(\boldsymbol{\alpha}=(-1.5,0,1.5)^\top\),
\(\boldsymbol{\beta}=(0.7,0.9)^\top\), and
\(\boldsymbol{\gamma}=(0,0.2,-1)^\top\). All GPPOM candidates are fitted.

```{r scenario-a, warning=FALSE}
scenario_a <- list(
  id = "a",
  true_Q_function = function(dat) {
    true_Q_ppom(dat)
  },
  fit_all_gppoms = TRUE
)

results_a <- run_simulation(
  scenario = scenario_a,
  R = n_rep,
  n = n_obs,
  m = m_cat,
  seed_start = base_seed
)

results_a$failed_rep
table(results_a$fit$model)
```

## Scenario (b): location-shift model

Data are generated from the location-shift model with
\(\boldsymbol{\alpha}=(-1.5,0,1.5)^\top\),
\(\boldsymbol{\beta}=(0.7,0.4)^\top\), and
\(\boldsymbol{\gamma}=(0.5,0,-0.5)^\top\). All GPPOM candidates are fitted.

```{r scenario-b, warning=FALSE}
scenario_b <- list(
  id = "b",
  true_Q_function = function(dat) {
    true_Q_location_shift(dat)
  },
  fit_all_gppoms = TRUE
)

results_b <- run_simulation(
  scenario = scenario_b,
  R = n_rep,
  n = n_obs,
  m = m_cat,
  seed_start = base_seed
)

results_b$failed_rep
table(results_b$fit$model)
```

## Scenario (c): GPPOM with gamma = (0, 0.2, -1)

Data are generated from the GPPOM specified by
\(S_2\{\rho_{1,1}(\mathbf{x};j)\}\),
\(\boldsymbol{\beta}=(0.7,0.9)^\top\), and
\(\boldsymbol{\gamma}=(0,0.2,-1)^\top\). The corresponding GPPOM is fitted.

```{r scenario-c, warning=FALSE}
scenario_c <- list(
  id = "c",
  true_Q_function = function(dat) {
    true_Q_gppom(
      dat,
      gamma = c(0, 0.2, -1)
    )
  },
  fit_all_gppoms = FALSE,
  gppom_surv = "S2",
  gppom_rhoj = "rho_11"
)

results_c <- run_simulation(
  scenario = scenario_c,
  R = n_rep,
  n = n_obs,
  m = m_cat,
  seed_start = base_seed
)

results_c$failed_rep
table(results_c$fit$model)
```

## Scenario (d): GPPOM with gamma = (0, -0.5, -1)

Data are generated from the GPPOM specified by
\(S_2\{\rho_{1,1}(\mathbf{x};j)\}\),
\(\boldsymbol{\beta}=(0.7,0.9)^\top\), and
\(\boldsymbol{\gamma}=(0,-0.5,-1)^\top\). The corresponding GPPOM is fitted.

```{r scenario-d, warning=FALSE}
scenario_d <- list(
  id = "d",
  true_Q_function = function(dat) {
    true_Q_gppom(
      dat,
      gamma = c(0, -0.5, -1)
    )
  },
  fit_all_gppoms = FALSE,
  gppom_surv = "S2",
  gppom_rhoj = "rho_11"
)

results_d <- run_simulation(
  scenario = scenario_d,
  R = n_rep,
  n = n_obs,
  m = m_cat,
  seed_start = base_seed
)

results_d$failed_rep
table(results_d$fit$model)
```

# Selection frequencies and Figure S1

```{r summary-and-figure-functions}
model_order <- c(
  "PPOM",
  "Location_shift",
  "GPPOM"
)

model_labels <- c(
  "PPOM" = "PPOM",
  "Location_shift" = "Location-shift",
  "GPPOM" = "GPPOM"
)

box_style <- list(
  width = 0.7,
  fill = "white",
  colour = "black",
  box.linewidth = 0.6,
  median.linewidth = 0.8,
  whisker.linetype = 2,
  whisker.linewidth = 0.5,
  staplewidth = 0.5,
  staple.linewidth = 0.5,
  outlier.shape = 1,
  outlier.size = 1.3,
  outlier.stroke = 0.45
)


# Prepare the three fitted models for tabulation and plotting.
prepare_comparison_data <- function(simulation_result) {
  simulation_result$fit %>%
    mutate(
      model = factor(model, levels = model_order),
      deviance = -2 * logLik
    )
}


# Calculate model-selection frequencies for AIC or BIC.
calculate_selection_frequency <- function(comparison_results,
                                          criterion = c("AIC", "BIC"),
                                          tolerance = 1e-8) {
  criterion <- match.arg(criterion)
  n_successful <- dplyr::n_distinct(comparison_results$rep)

  comparison_results %>%
    group_by(rep) %>%
    mutate(
      best_value = min(.data[[criterion]], na.rm = TRUE),
      is_best = abs(.data[[criterion]] - best_value) < tolerance
    ) %>%
    filter(is_best) %>%
    summarise(
      winner = paste(
        model_labels[as.character(model)],
        collapse = " / "
      ),
      .groups = "drop"
    ) %>%
    count(winner, name = "Frequency") %>%
    mutate(Percent = 100 * Frequency / n_successful) %>%
    arrange(desc(Frequency), winner)
}

# Calculate the medians represented by the boxplot center lines.
calculate_boxplot_medians <- function(comparison_results) {
  comparison_results %>%
    group_by(model) %>%
    summarise(
      median_deviance = median(deviance, na.rm = TRUE),
      median_MSE = median(MSE_Q, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      model = model_labels[as.character(model)]
    ) %>%
    as.data.frame()
}

# Create the deviance and MSE panels for one scenario.
make_plot_pair <- function(comparison_results) {
  common_theme <- theme_bw(base_size = 12) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(
        colour = "black",
        fill = NA,
        linewidth = 0.4
      ),
      axis.text.x = element_text(
        angle = 0,
        hjust = 0.5,
        vjust = 1,
        size = 10
      ),
      axis.text.y = element_text(size = 10),
      axis.title.y = element_text(size = 12)
    )

  p_deviance <- ggplot(
    comparison_results,
    aes(x = model, y = deviance)
  ) +
    do.call(geom_boxplot, box_style) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "Deviance") +
    common_theme

  p_mse <- ggplot(
    comparison_results,
    aes(x = model, y = MSE_Q)
  ) +
    do.call(geom_boxplot, box_style) +
    scale_x_discrete(labels = model_labels) +
    labs(x = NULL, y = "MSE") +
    common_theme

  list(deviance = p_deviance, mse = p_mse)
}
```

## AIC and BIC selection frequencies

### Scenario (a)

```{r selection-frequencies-a}
comparison_a <- prepare_comparison_data(results_a)

AIC_frequency_a <- calculate_selection_frequency(
  comparison_a,
  criterion = "AIC"
)

BIC_frequency_a <- calculate_selection_frequency(
  comparison_a,
  criterion = "BIC"
)

AIC_frequency_a
BIC_frequency_a
```

### Scenario (b)

```{r selection-frequencies-b}
comparison_b <- prepare_comparison_data(results_b)

AIC_frequency_b <- calculate_selection_frequency(
  comparison_b,
  criterion = "AIC"
)

BIC_frequency_b <- calculate_selection_frequency(
  comparison_b,
  criterion = "BIC"
)

AIC_frequency_b
BIC_frequency_b
```

### Scenario (c)

```{r selection-frequencies-c}
comparison_c <- prepare_comparison_data(results_c)

AIC_frequency_c <- calculate_selection_frequency(
  comparison_c,
  criterion = "AIC"
)

BIC_frequency_c <- calculate_selection_frequency(
  comparison_c,
  criterion = "BIC"
)

AIC_frequency_c
BIC_frequency_c
```

### Scenario (d)

```{r selection-frequencies-d}
comparison_d <- prepare_comparison_data(results_d)

AIC_frequency_d <- calculate_selection_frequency(
  comparison_d,
  criterion = "AIC"
)

BIC_frequency_d <- calculate_selection_frequency(
  comparison_d,
  criterion = "BIC"
)

AIC_frequency_d
BIC_frequency_d
```

## Boxplot medians

### Scenario (a)

```{r boxplot-medians-a}
median_summary_a <- calculate_boxplot_medians(comparison_a)
median_summary_a
```

### Scenario (b)

```{r boxplot-medians-b}
median_summary_b <- calculate_boxplot_medians(comparison_b)
median_summary_b
```

### Scenario (c)

```{r boxplot-medians-c}
median_summary_c <- calculate_boxplot_medians(comparison_c)
median_summary_c
```

### Scenario (d)

```{r boxplot-medians-d}
median_summary_d <- calculate_boxplot_medians(comparison_d)
median_summary_d
```

## Figure S1

```{r figure-s1, fig.width=14, fig.height=16}
plots_a <- make_plot_pair(comparison_a)
plots_b <- make_plot_pair(comparison_b)
plots_c <- make_plot_pair(comparison_c)
plots_d <- make_plot_pair(comparison_d)

p_deviance_a <- plots_a$deviance +
  labs(title = "(a) PPOM") +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 14
    )
  )

p_deviance_b <- plots_b$deviance +
  labs(title = "(b) Location-shift") +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 14
    )
  )

p_deviance_c <- plots_c$deviance +
  labs(
    title = expression(
      bold("(c) GPPOM with") ~ S[2](rho[1*','*1](x*';'*j))
    ),
    subtitle = expression(
      bold(gamma) == group("(", list(0, 0.2, -1), ")")^T
    )
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 14
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      face = "bold",
      size = 11
    )
  )

p_deviance_d <- plots_d$deviance +
  labs(
    title = expression(
      bold("(d) GPPOM with") ~ S[2](rho[1*','*1](x*';'*j))
    ),
    subtitle = expression(
      bold(gamma) == group("(", list(0, -0.5, -1), ")")^T
    )
  ) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 14
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      face = "bold",
      size = 11
    )
  )

figure_s1 <-
  (p_deviance_a | p_deviance_b) /
  (plots_a$mse | plots_b$mse) /
  (p_deviance_c | p_deviance_d) /
  (plots_c$mse | plots_d$mse) +
  plot_layout(
    widths = c(1, 1),
    heights = c(1, 1, 1, 1)
  )

figure_s1
```
