---
title: "APS data analysis: Table 5"
output:
  html_document:
    code_folding: show
    highlight: textmate
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

This file reproduces Table 5 in Section 4.8. The coefficient estimates,
standard errors, log-likelihoods, AIC values, and BIC values are obtained from
fits to the complete APS data. The LogS and RPS values are obtained by repeated
stratified 10-fold cross-validation.

```{r setup, message=FALSE}
library(ordinal)
library(zipfR)

source(file.path("code", "common_functions.R"))
```

# Data

```{r data}
data("aps", package = "aplore3")

aps$danger <- factor(
  aps$danger,
  levels = c("Unlikely", "Possible", "Probable", "Likely"),
  ordered = TRUE
)
aps$weeks <- aps$los / 7

analysis_variables <- c("danger", "weeks", "behav", "gender", "custd")
aps_analysis <- aps[, analysis_variables, drop = FALSE]

if (anyNA(aps_analysis)) {
  stop("Missing values are present in the APS analysis data.")
}

aps_analysis$.row_id <- seq_len(nrow(aps_analysis))
outcome_levels <- levels(aps_analysis$danger)
aps_formula <- danger ~ weeks + behav + gender + custd
```

# Common GCLM functions

The three nonidentifiable combinations under the continuous-covariate setting
are excluded from both the complete-data analysis and cross-validation.

```{r gclm-functions}
is_identifiable_gclm <- function(surv_name, rhoj_name) {
  !(surv_name == "H4" &&
      rhoj_name %in% c("rho_14", "rho_44", "rho_54"))
}

fit_all_gclm_candidates <- function(X, y, m, show_progress = FALSE) {
  surv_candidates <- make_surv_func(m = m)
  rhoj_candidates <- make_rhoj_candidates()
  all_results <- list()

  for (surv_name in names(surv_candidates)) {
    for (rhoj_name in names(rhoj_candidates)) {
      if (!is_identifiable_gclm(surv_name, rhoj_name)) {
        next
      }

      eta_name <- get_eta_name_for_rhoj(rhoj_name)
      surv_info <- surv_candidates[[surv_name]]
      eta_info <- geta_candidates[[eta_name]]
      rhoj_info <- rhoj_candidates[[rhoj_name]]

      init_fit <- get_data_driven_init(
        y = y,
        m = m,
        surv_info = surv_info,
        eta_info = eta_info,
        rhoj_info = rhoj_info
      )

      surv_info$init <- init_fit$surv_init
      rhoj_info$init <- init_fit$rhoj_init

      fit <- fit_one_combo(
        X = X,
        y = y,
        m = m,
        n = nrow(X),
        surv_info = surv_info,
        eta_info = eta_info,
        rhoj_info = rhoj_info,
        method = "BFGS"
      )

      combination_key <- paste(surv_name, eta_name, rhoj_name, sep = "__")
      all_results[[combination_key]] <- c(
        list(
          surv = surv_name,
          geta = eta_name,
          rhoj = rhoj_name
        ),
        fit
      )

      if (show_progress) {
        message("Finished: ", combination_key)
      }
    }
  }

  all_results
}

select_best_gclm <- function(all_results, criterion = c("BIC", "AIC")) {
  criterion <- match.arg(criterion)
  criterion_field <- tolower(criterion)

  criterion_values <- vapply(
    all_results,
    function(fit) {
      value <- fit[[criterion_field]]
      eligible <-
        length(fit$conv) == 1L &&
        fit$conv == 0L &&
        length(value) == 1L &&
        is.finite(value)

      if (isTRUE(eligible)) value else Inf
    },
    numeric(1)
  )

  if (all(!is.finite(criterion_values))) {
    stop("No identifiable GCLM candidate converged with a finite ", criterion, ".")
  }

  selected_key <- names(criterion_values)[which.min(criterion_values)]

  list(
    key = selected_key,
    fit = all_results[[selected_key]]
  )
}

predict_gclm_probabilities <- function(X, fit, m) {
  surv_candidates <- make_surv_func(m = m)
  rhoj_candidates <- make_rhoj_candidates()

  eta <- drop(X %*% fit$beta)
  eta_value <- geta_candidates[[fit$geta]]$func(eta)

  conditional_survival <- vapply(
    seq_len(m - 1L),
    function(j) {
      rho_j <- do.call(
        rhoj_candidates[[fit$rhoj]]$fun,
        c(
          list(eta_val = eta_value, j = j, m = m),
          as.list(fit$rhoj_par)
        )
      )

      do.call(
        surv_candidates[[fit$surv]]$func,
        c(list(t = rho_j), as.list(fit$surv_par))
      )
    },
    FUN.VALUE = numeric(nrow(X))
  )

  conditional_survival <- matrix(
    conditional_survival,
    nrow = nrow(X),
    ncol = m - 1L
  )

  cbind(1, conditional_survival) - cbind(conditional_survival, 0)
}
```

# Complete-data fits

```{r complete-data-fits, warning=FALSE}
clm_fits <- list(
  Logit = ordinal::clm(aps_formula, data = aps_analysis, link = "logit"),
  Probit = ordinal::clm(aps_formula, data = aps_analysis, link = "probit"),
  Cloglog = ordinal::clm(aps_formula, data = aps_analysis, link = "cloglog")
)

X_complete <- model.matrix(
  ~ weeks + behav + gender + custd,
  data = aps_analysis
)[, -1, drop = FALSE]
y_complete <- as.integer(aps_analysis$danger)
m <- length(outcome_levels)

gclm_complete_results <- fit_all_gclm_candidates(
  X = X_complete,
  y = y_complete,
  m = m,
  show_progress = FALSE
)

best_gclm_complete <- select_best_gclm(
  gclm_complete_results,
  criterion = "BIC"
)

```

```{r complete-data-summary}
complete_data_summary <- lapply(
  names(clm_fits),
  function(model_name) {
    fit <- clm_fits[[model_name]]
    coefficient_table <- coef(summary(fit))
    coefficient_rows <- seq.int(
      nrow(coefficient_table) - ncol(X_complete) + 1L,
      nrow(coefficient_table)
    )
    coefficient_table <- coefficient_table[
      coefficient_rows,
      ,
      drop = FALSE
    ]

    list(
      estimate = coefficient_table[, "Estimate"],
      standard_error = coefficient_table[, "Std. Error"],
      number_of_parameters = as.integer(attr(logLik(fit), "df")),
      logLik = as.numeric(logLik(fit)),
      AIC = AIC(fit),
      BIC = BIC(fit)
    )
  }
)
names(complete_data_summary) <- names(clm_fits)

selected_gclm_fit <- best_gclm_complete$fit

complete_data_summary$GCLM <- list(
  estimate = selected_gclm_fit$beta,
  standard_error = selected_gclm_fit$se_beta,
  number_of_parameters = length(selected_gclm_fit$par),
  logLik = -selected_gclm_fit$nll,
  AIC = selected_gclm_fit$aic,
  BIC = selected_gclm_fit$bic
)
```

# Repeated stratified 10-fold cross-validation

No covariate preprocessing or standardization is applied in this analysis.

```{r cv-functions}
make_stratified_folds <- function(y, v = 10L, repeats = 10L, seed = 1234L) {
  y <- droplevels(factor(y))
  category_counts <- table(y)

  if (any(category_counts < v)) {
    stop(
      "Every response category must contain at least v observations. Counts: ",
      paste(names(category_counts), category_counts, sep = "=", collapse = ", ")
    )
  }

  set.seed(seed)

  fold_list <- lapply(seq_len(repeats), function(repeat_id) {
    fold_id <- integer(length(y))

    for (level in levels(y)) {
      row_index <- which(y == level)
      balanced_labels <- rep(seq_len(v), length.out = length(row_index))
      fold_id[row_index] <- sample(
        balanced_labels,
        length(row_index),
        replace = FALSE
      )
    }

    data.frame(
      repeat_id = repeat_id,
      fold_id = fold_id,
      row_id = seq_along(y)
    )
  })

  folds <- do.call(rbind, fold_list)
  rownames(folds) <- NULL
  folds
}

validate_probabilities <- function(prob, n, outcome_levels, tolerance = 1e-7) {
  prob <- as.matrix(prob)

  if (nrow(prob) != n || ncol(prob) != length(outcome_levels)) {
    stop("The prediction must be an n by m matrix of category probabilities.")
  }

  if (!is.null(colnames(prob))) {
    if (!setequal(colnames(prob), outcome_levels)) {
      stop("Prediction column names do not match the outcome levels.")
    }
    prob <- prob[, outcome_levels, drop = FALSE]
  }

  if (any(!is.finite(prob))) {
    stop("Non-finite predicted probabilities were returned.")
  }
  if (any(prob < -tolerance) || any(prob > 1 + tolerance)) {
    stop("Predicted probabilities lie outside [0, 1].")
  }
  if (any(abs(rowSums(prob) - 1) > tolerance)) {
    stop("Predicted category probabilities do not sum to one.")
  }

  prob[prob < 0] <- 0
  prob[prob > 1] <- 1
  prob <- prob / rowSums(prob)
  colnames(prob) <- outcome_levels
  prob
}

calculate_observation_scores <- function(truth, prob, outcome_levels) {
  truth <- factor(truth, levels = outcome_levels, ordered = TRUE)
  outcome_index <- as.integer(truth)
  n <- length(truth)
  m <- length(outcome_levels)

  if (anyNA(outcome_index)) {
    stop("The observed outcome contains an unknown category.")
  }

  observed_probability <- prob[cbind(seq_len(n), outcome_index)]
  log_score <- -log(observed_probability)

  cumulative_probability <- t(apply(prob, 1L, cumsum))[
    , seq_len(m - 1L), drop = FALSE
  ]
  observed_cumulative <- outer(
    outcome_index,
    seq_len(m - 1L),
    FUN = "<="
  )
  ranked_probability_score <- rowSums(
    (cumulative_probability - observed_cumulative)^2
  )

  data.frame(
    LogS = log_score,
    RPS = ranked_probability_score
  )
}

fit_predict_clm <- function(train, test, outcome_levels, link) {
  fit <- ordinal::clm(
    aps_formula,
    data = train,
    link = link
  )

  newdata <- test
  newdata$danger <- NULL

  probability <- predict(
    fit,
    newdata = newdata,
    type = "prob"
  )$fit

  probability <- as.matrix(probability)
  colnames(probability) <- outcome_levels

  list(
    prob = probability,
    selected = NA_character_
  )
}

fit_predict_gclm <- function(train, test, outcome_levels) {
  y_train <- match(as.character(train$danger), outcome_levels)
  m <- length(outcome_levels)

  if (anyNA(y_train)) {
    stop("Unknown response category in the training data.")
  }

  X_train <- model.matrix(
    ~ weeks + behav + gender + custd,
    data = train
  )[, -1, drop = FALSE]
  X_test <- model.matrix(
    ~ weeks + behav + gender + custd,
    data = test
  )[, -1, drop = FALSE]

  if (!setequal(colnames(X_train), colnames(X_test))) {
    stop("Training and test design matrices have different columns.")
  }
  X_test <- X_test[, colnames(X_train), drop = FALSE]

  all_results <- fit_all_gclm_candidates(
    X = X_train,
    y = y_train,
    m = m,
    show_progress = FALSE
  )
  selected <- select_best_gclm(all_results, criterion = "BIC")

  probability <- predict_gclm_probabilities(
    X = X_test,
    fit = selected$fit,
    m = m
  )
  colnames(probability) <- outcome_levels

  list(prob = probability, selected = selected$key)
}

model_specs <- list(
  Logit = function(train, test, outcome_levels) {
    fit_predict_clm(train, test, outcome_levels, link = "logit")
  },
  Probit = function(train, test, outcome_levels) {
    fit_predict_clm(train, test, outcome_levels, link = "probit")
  },
  Cloglog = function(train, test, outcome_levels) {
    fit_predict_clm(train, test, outcome_levels, link = "cloglog")
  },
  GCLM = fit_predict_gclm
)

run_repeated_cv <- function(data, folds, model_specs, outcome = "danger") {
  outcome_levels <- levels(data[[outcome]])
  repeat_values <- sort(unique(folds$repeat_id))
  score_results <- list()
  selection_results <- list()
  failure_results <- list()
  score_counter <- 0L
  selection_counter <- 0L
  failure_counter <- 0L

  for (repeat_id in repeat_values) {
    fold_values <- sort(unique(folds$fold_id[folds$repeat_id == repeat_id]))

    for (fold_id in fold_values) {
      message(
        sprintf(
          "CV repeat %d/%d, fold %d/%d",
          match(repeat_id, repeat_values),
          length(repeat_values),
          match(fold_id, fold_values),
          length(fold_values)
        )
      )

      test_ids <- folds$row_id[
        folds$repeat_id == repeat_id & folds$fold_id == fold_id
      ]
      train <- data[!data$.row_id %in% test_ids, , drop = FALSE]
      test <- data[data$.row_id %in% test_ids, , drop = FALSE]

      for (model_name in names(model_specs)) {
        result <- tryCatch(
          model_specs[[model_name]](train, test, outcome_levels),
          error = function(error) error
        )

        if (inherits(result, "error")) {
          failure_counter <- failure_counter + 1L
          failure_results[[failure_counter]] <- data.frame(
            repeat_id = repeat_id,
            fold_id = fold_id,
            model = model_name,
            message = conditionMessage(result)
          )
          next
        }

        probability <- tryCatch(
          validate_probabilities(
            result$prob,
            n = nrow(test),
            outcome_levels = outcome_levels
          ),
          error = function(error) error
        )

        if (inherits(probability, "error")) {
          failure_counter <- failure_counter + 1L
          failure_results[[failure_counter]] <- data.frame(
            repeat_id = repeat_id,
            fold_id = fold_id,
            model = model_name,
            message = conditionMessage(probability)
          )
          next
        }

        observation_scores <- calculate_observation_scores(
          truth = test[[outcome]],
          prob = probability,
          outcome_levels = outcome_levels
        )

        score_counter <- score_counter + 1L
        score_results[[score_counter]] <- data.frame(
          repeat_id = repeat_id,
          fold_id = fold_id,
          row_id = test$.row_id,
          model = model_name,
          observation_scores
        )

        if (model_name == "GCLM") {
          selection_counter <- selection_counter + 1L
          selection_results[[selection_counter]] <- data.frame(
            repeat_id = repeat_id,
            fold_id = fold_id,
            selected = result$selected
          )
        }
      }
    }
  }

  scores <- if (length(score_results)) {
    do.call(rbind, score_results)
  } else {
    data.frame()
  }
  selections <- if (length(selection_results)) {
    do.call(rbind, selection_results)
  } else {
    data.frame()
  }
  failures <- if (length(failure_results)) {
    do.call(rbind, failure_results)
  } else {
    data.frame(
      repeat_id = integer(),
      fold_id = integer(),
      model = character(),
      message = character()
    )
  }

  list(scores = scores, selections = selections, failures = failures)
}

summarise_cv_scores <- function(cv_result) {
  if (!nrow(cv_result$scores)) {
    stop("There are no successful held-out predictions.")
  }
  if (nrow(cv_result$failures) > 0L) {
    print(cv_result$failures)
    stop("Some model-fold fits failed.")
  }

  by_repeat <- aggregate(
    cbind(LogS, RPS) ~ model + repeat_id,
    data = cv_result$scores,
    FUN = mean
  )
  overall <- aggregate(
    cbind(LogS, RPS) ~ model,
    data = by_repeat,
    FUN = mean
  )

  overall[match(names(model_specs), overall$model), , drop = FALSE]
}
```

```{r run-cv, warning=FALSE}
folds <- make_stratified_folds(
  y = aps_analysis$danger,
  v = 10L,
  repeats = 10L,
  seed = 1234L
)

cv_result <- run_repeated_cv(
  data = aps_analysis,
  folds = folds,
  model_specs = model_specs,
  outcome = "danger"
)

cv_summary <- summarise_cv_scores(cv_result)
```

# Display the results reported in Table 5

```{r table-5}
format_estimate <- function(estimate, standard_error) {
  sprintf("%.3f (%.3f)", estimate, standard_error)
}

model_order <- c("Logit", "Probit", "Cloglog", "GCLM")

table5_results <- data.frame(
  `Param.` = c(
    "Weeks",
    "Behav",
    "Gender",
    "Custd",
    "# of param.",
    "logLik",
    "AIC",
    "BIC",
    "LogS",
    "RPS"
  ),
  check.names = FALSE
)

for (model_name in model_order) {
  complete_summary <- complete_data_summary[[model_name]]
  cv_row <- cv_summary[cv_summary$model == model_name, , drop = FALSE]

  table5_results[[model_name]] <- c(
    format_estimate(
      complete_summary$estimate,
      complete_summary$standard_error
    ),
    sprintf("%d", complete_summary$number_of_parameters),
    sprintf("%.3f", complete_summary$logLik),
    sprintf("%.3f", complete_summary$AIC),
    sprintf("%.3f", complete_summary$BIC),
    sprintf("%.3f", cv_row$LogS),
    sprintf("%.3f", cv_row$RPS)
  )
}

print(table5_results, row.names = FALSE, right = FALSE)
```
